---
title: "Dissertation Statistical Analysis"
author: "B286413"
date: "2026-06-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Downloading packages
# BioManager
#install.packages("BiocManager", repos = "https://cloud.r-project.org")
# phyloseq
#BiocManager::install("phyloseq", ask = FALSE, update = FALSE)
# QIIME2
#install.packages("remotes", repos = "https://cloud.r-project.org")
#remotes::install_github("jbisanz/qiime2R")
# picante
#BiocManager::install("picante", ask = FALSE, update = FALSE)
# ggpubr
#install.packages("ggpubr")
# microbiome
#BiocManager::install("microbiome")
# lmerTest
#install.packages("lmerTest")
#DHARMa
#install.packages("DHARMa")
# Masslin2
#BiocManager::install("Maaslin2")
# MaAsLin 3
#BiocManager::install("biobakery/maaslin3")
# mia
#BiocManager::install("mia")
# pairwiseAdonis
#remotes::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

# Loading packages
library(tidyverse)
library(phyloseq)
library(qiime2R)
library(vegan)
library(picante)
library(ggpubr)
library(microbiome)
library(lme4)
library(MASS)
library(DHARMa)
library(Maaslin2)
library(maaslin3)
library(mia)
library(emmeans)
library(pairwiseAdonis)

# Resolving conflicts
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::count)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::desc)
conflicts_prefer(base::intersect)
conflicts_prefer(dplyr::inner_join)
conflicts_prefer(base::summary)
conflicts_prefer(dplyr::left_join)
conflicts_prefer(dplyr::rename)
conflicts_prefer(base::as.factor)

# Working directory
setwd("~/project")
```

## Loading Data

### Data Description

The data were generated following a standard QIIME2 analysis workflow. Files are
located in the following relative paths:
* ./denoise_set2_short/merged_asv_table.qza - merged feature table
* ./metadata/metadata_clean.rds - clean metadata file
* ./taxonomy_set2_short/merged_taxonomy.qza - taxonomic classification file
* ./phylogeny_set2_short/rooted_tree.qza - rooted phylogenetic tree

```{r loading_data}
## Paths
path_table <- "./denoise_set2_short/merged_asv_table.qza"
path_tax <- "./taxonomy_set2_short/merged_taxonomy.qza"
path_tree <- "./phylogeny_set2_short/rooted_tree.qza"
path_md_rds <- "./metadata/metadata_clean.rds"

# Feature table
features <- read_qza(path_table)$data
colnames(features) <- trimws(as.character(colnames(features)))

# Metadata
md <- readRDS(path_md_rds)
md <- as.data.frame(md)

# Cleaning sample IDs
colnames(features) <- trimws(as.character(colnames(features)))
md$activation_code <- trimws(as.character(md$activation_code))
md <- md[!is.na(md$activation_code) & md$activation_code != "", , drop = FALSE]
rownames(md) <- md$activation_code

# Checking duplicates
stopifnot(anyDuplicated(colnames(features)) == 0)
stopifnot(anyDuplicated(rownames(md)) == 0)

# Subsetting to metadata overlap, keeping feature-table order
keep_samples <- colnames(features)[colnames(features) %in% rownames(md)]
features <- features[, keep_samples, drop = FALSE]

# Reordering metadata to match features columns
md <- md[colnames(features), , drop = FALSE]

# Sanity check
stopifnot(identical(colnames(features), rownames(md)))

# Removing absent taxa after subsetting
keep_taxa_table <- rownames(features)[rowSums(features) > 0]
features <- features[keep_taxa_table, , drop = FALSE]

# Taxonomy
tax_df <- read_qza(path_tax)$data
# Updating row names
rownames(tax_df) <- as.character(tax_df$Feature.ID)
# Converting taxonomy into a rank matrix for phyloseq
tax_mat <- str_split_fixed(as.character(tax_df$Taxon), ";\\s*", n = 7)
colnames(tax_mat) <- c("Domain", "Phylum", "Class", "Order",
                       "Family", "Genus", "Species")
rownames(tax_mat) <- rownames(tax_df)

# Phylogenetic tree
tree_phylo <- read_qza(path_tree)$data

# Intersection across all components
common_taxa <- Reduce(intersect, list(
  rownames(features),
  rownames(tax_mat),
  tree_phylo$tip.label
))

# Pruning all components to shared taxa
features2 <- features[common_taxa, , drop = FALSE]
tax_mat2 <- tax_mat[common_taxa, , drop = FALSE]
tree2_phylo <- ape::keep.tip(tree_phylo, common_taxa)

# phyloseq components
otu <- otu_table(as.matrix(features2), taxa_are_rows = TRUE)
tax <- tax_table(tax_mat2)
tr <- phy_tree(tree2_phylo)
smd <- sample_data(md)

# phyloseq object
ps <- phyloseq(otu, tax, tr, smd)

# Verifying object - sanity checks
ps
validObject(ps)
nsamples(ps)
ntaxa(ps)

# Depth statistics
summary(sample_sums(ps))
summary(taxa_sums(ps))

# Zero depth samples
sum(sample_sums(ps) == 0)

# Initial depth vs. richness inspection
sample_data(ps)$depth <- sample_sums(ps)
p <- plot_richness(ps, x = "depth", measures = "Observed") +
  scale_x_log10() +
  ggtitle("Observed Richness vs. Depth (pre-filtering)") +
  ylab("Observed Richness") +
  xlab("Sample Depth") +
  theme_bw() +
  theme(
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
p

# Lowest-depth samples
sort(sample_sums(ps))[1:20]

# Highest richness samples
alpha <- estimate_richness(ps, measures = "Observed")
rownames(alpha)[order(alpha$Observed, decreasing = TRUE)][1:20]
alpha[order(alpha$Observed, decreasing = TRUE),
      "Observed", drop = FALSE][1:20, ]
```

## Batch Effect Exploration

The hundreds of samples come from different sequencing batches and DNA
extracting batches. Samples from the same batch were not collected at the same
time, so there could also be storage effect. The available data contain
information about sequencing batches only, which was explored fora potential
effect. This exploration is not included in the final analysis. 

```{r batch_effect, eval=FALSE}
# Verifying batch is a factor
sample_data(ps)$batch <- as.factor(sample_data(ps)$batch)

# Adding reads and alpha diversity metrics into sample_data
sample_data(ps)$depth <- sample_sums(ps)
alpha <- estimate_richness(ps,
                           measures = c("Observed", "Shannon", "InvSimpson"))
sample_data(ps)$observed <- alpha[sample_names(ps), "Observed"]
sample_data(ps)$shannon  <- alpha[sample_names(ps), "Shannon"]
sample_data(ps)$simpson  <- alpha[sample_names(ps), "InvSimpson"]

# Faith's PD
otu_m <- as.matrix(otu_table(ps))
if (taxa_are_rows(ps)) otu_m <- t(otu_m)
pd_df <- picante::pd(otu_m, phy_tree(ps))
faith_pd <- pd_df$PD
names(faith_pd) <- rownames(pd_df)
sample_data(ps)$faith_pd <- faith_pd[sample_names(ps)]

# Single plotting table
meta_batch <- data.frame(sample_data(ps), stringsAsFactors = FALSE) %>%
  rownames_to_column("sample_id") %>%
  filter(!is.na(batch)) %>%
  mutate(batch = factor(batch))

# Reads by batch
batch_reads <- ggplot(meta_batch, aes(x = batch, y = depth, fill = batch)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "GnBu") +
  stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  ggtitle("Number of Reads by Sequencing Batch") +
  ylab("Number of Reads") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  )

# Summary statistics by batch
reads_summary_batch <- meta_batch %>%
  group_by(batch) %>%
  summarize(
    n = n(),
    min = min(depth, na.rm = TRUE),
    max = max(depth, na.rm = TRUE),
    median = median(depth, na.rm = TRUE),
    mean = mean(depth, na.rm = TRUE),
    .groups = "drop"
  )

# Alpha diversity boxplots by batch
batch_alpha <- ggarrange(
  # Observed ASVs
  ggplot(meta_batch, aes(x = batch, y = observed, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Shannon's Index
  ggplot(meta_batch, aes(x = batch, y = shannon, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(meta_batch, aes(x = batch, y = simpson, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(meta_batch, aes(x = batch, y = faith_pd, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),
  ncol = 3, nrow = 1,
  common.legend = TRUE, legend = "bottom", align = "hv"
)

batch_alpha <- annotate_figure(batch_alpha,
                               top =
                                 text_grob("Alpha Diversity Metrics per Batch",
                                           face = "bold", size = 14))

# Alpha diversity summary statistics by batch
alpha_summary_batch <- meta_batch %>%
  group_by(batch) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(simpson, na.rm = TRUE),
    max_simpson = max(simpson, na.rm = TRUE),
    median_simpson = median(simpson, na.rm = TRUE),
    mean_simpson = mean(simpson, na.rm = TRUE),
    min_faith = min(faith_pd, na.rm = TRUE),
    max_faith = max(faith_pd, na.rm = TRUE),
    median_faith = median(faith_pd, na.rm = TRUE),
    mean_faith = mean(faith_pd, na.rm = TRUE),
    .groups = "drop"
  )

# Beta diversity and PCoA by batch
dist_bray <- phyloseq::distance(ps, method = "bray")
dist_jac <- phyloseq::distance(ps, method = "jaccard", binary = TRUE)

# Previous pruning could have led to a non-dichotomic tree
# https://github.com/joey711/phyloseq/issues/936
phy_tree(ps) <- ape::multi2di(phy_tree(ps))
dist_uwu <- phyloseq::UniFrac(ps, weighted = FALSE, normalized = TRUE)
dist_wu <- phyloseq::UniFrac(ps, weighted = TRUE,  normalized = TRUE)

# Ordinations
set.seed(122863)
ord_bray <- ordinate(ps, method = "PCoA", distance = dist_bray)
set.seed(122863)
ord_jac <- ordinate(ps, method = "PCoA", distance = dist_jac)
set.seed(122863)
ord_uwu <- ordinate(ps, method = "PCoA", distance = dist_uwu)
set.seed(122863)
ord_wu <- ordinate(ps, method = "PCoA", distance = dist_wu)

batch_beta <- ggarrange(
  plot_ordination(ps, ord_bray, color = "batch") +
    geom_point(size = 2) +
    labs(color = "Sequencing batch", caption = "Bray-Curtis PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  plot_ordination(ps, ord_jac, color = "batch") +
    geom_point(size = 2) +
    labs(color = "Sequencing batch", caption = "Jaccard (binary) PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  plot_ordination(ps, ord_uwu, color = "batch") +
    geom_point(size = 2) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Sequencing batch", caption = "Unweighted UniFrac PCoA"),

  plot_ordination(ps, ord_wu, color = "batch") +
    geom_point(size = 2) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Sequencing batch", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
batch_beta <- annotate_figure(batch_beta,
                              top =
                                text_grob("Beta Diversity Metrics per Batch",
                                          face = "bold", size = 14))
```

## Filtering Data

Initial inspection in QIIME2 revealed that the raw data contains mitochondrial
and chloroplast taxa, which should be removed, along with eukaryotic sequences.
This part will also remove irregular samples that were previously identified.

```{r filtering}
# Pre-filtering object dimensions
nsamples(ps) # 1063
ntaxa(ps)
summary(sample_sums(ps))

# Metadata
md_f <- as(sample_data(ps), "data.frame")

# Duplicated samples based on name and dob -------------------------------------
# Nine in total, two are in the May batch
dup_keys <- md_f %>%
  count(name, dob, name = "n") %>%
  filter(n > 1) %>%
  arrange(desc(n))
dup_keys

# Duplicated rows for metadata inspection
dup_rows <- md_f %>%
  inner_join(dup_keys, by = c("name", "dob")) %>%
  arrange(name, dob)
dup_rows

# Duplicated samples to remove based on metadata
dups <- c(
  "TW59F9", # Almost identical, fewer samples in the other batch
  "KHG2HC", # Fewer samples in other batch
  "VVYVBT", # Low vs. medium physical activity
  "B9KPQR", # Fewer samples in other batch
#  "REUBN8", # Low vs. medium physical activity - May batch
  "ZCJCAR", # No preference, picked the first one
  "P93VRB", # Overweight vs. normalweight
  "D2WG8P" # Raw vs. mixed diet
#  "JKXWMG" # Weight - the other one is in the May batch
)

# Filtering out "bad" samples --------------------------------------------------
# Missing metadata (NA, e.g., negative age) - four samples
missing_md <- unlist(sample_data(ps)[apply(
  sample_data(ps),
  1,
  function(x) any(is.na(x))
), "activation_code"])

# Irregular samples (writing all, some were already filtered in previous steps)
bad_samples <- c(
  dups,
  missing_md,
  "GER69C", # Irregularly low depth
  "86PGKK", # Irregularly low depth
  "S4F64A", # Low depth - under 40k
  "NVTNX3", # Low depth - under 40k
  "PGZJAP", # Low depth - under 40k
  "E27ADN" # Low depth - under 40k
)

# Removing samples and verifying consistent remainder
ps_f <- prune_samples(setdiff(sample_names(ps), bad_samples), ps)
nsamples(ps_f) # 1047

# Chloroplast, mitochondria and Eukaryota filtering ----------------------------
# Original occurrences
sum(grepl("mitochondria", as.vector(tax_table(ps_f)[, "Family"]),
          ignore.case = TRUE), na.rm = TRUE) # 50 matches
sum(grepl("chloroplast", as.vector(tax_table(ps_f)[, "Order"]),
          ignore.case = TRUE), na.rm = TRUE) # 59 matches
sum(grepl("eukaryota", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 182 matches, 3 times more
sum(grepl("Unassigned", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 184

# Checking if these are concentrated in particular samples
contam_taxa <- taxa_names(ps_f)[
  grepl("eukaryota",
        as.vector(tax_table(ps_f)[, "Domain"]),
        ignore.case = TRUE) |
    grepl("chloroplast",
          as.vector(tax_table(ps_f)[, "Order"]),
          ignore.case = TRUE) |
    grepl("mitochondria",
          as.vector(tax_table(ps_f)[, "Family"]),
          ignore.case = TRUE)
]

ps_contam <- prune_taxa(contam_taxa, ps_f)

# Per-sample summary: contaminant reads, total reads, and proportion
per_sample <- tibble(
  sample = sample_names(ps_f),
  contam_reads = sample_sums(ps_contam)[sample_names(ps_f)],
  total_reads = sample_sums(ps_f),
) %>%
  mutate(contam_prop = contam_reads / total_reads,
         contam_pct  = 100 * contam_prop) %>%
  arrange(desc(contam_pct))

# Worst samples
head(per_sample, 20)
summary(per_sample$contam_pct)

# Some more Eukaryota investigation --------------------------------------------
# Percentage
sum(grepl("eukaryota", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) * 100 / ntaxa(ps_f) # 1.439%

# Singletons
euk_taxa <- taxa_names(ps_f)[grepl("eukaryota",
                                   as.vector(tax_table(ps_f)[, "Domain"]),
                                   ignore.case = TRUE)]
euk_sums <- taxa_sums(prune_taxa(euk_taxa, ps_f))
# 0-24, mean 2.901, median 2
summary(euk_sums)
sum(euk_sums == 1) / length(euk_sums) # No singletons

# Per-batch inspection
ps_euk <- prune_taxa(euk_taxa, ps_f)

otu_euk <- as(otu_table(ps_euk), "matrix")
if (taxa_are_rows(ps_euk)) {
  otu_euk <- t(otu_euk)
}
batch <- as.character(sample_data(ps_f)$batch)
euk_batch <- data.frame(sample = sample_names(ps_f), batch,
                        euk_reads = rowSums(otu_euk),
                        total = sample_sums(ps_f),
                        euk_feats = rowSums(otu_euk > 0)) %>%
  group_by(batch) %>%
  summarize(n = n(),
            euk_pct = 100 * sum(euk_reads) / sum(total),
            mean_euk_feats = mean(euk_feats),
            .groups = "drop") %>%
  arrange(desc(euk_pct))
euk_batch

# Filtering based on taxa ------------------------------------------------------
# High content of eukaryotic sequences in 21_May_2026, excluding batch
ps_f <- subset_samples(ps_f, batch != "21-05-26")
ps_f <- prune_taxa(taxa_sums(ps_f) > 0, ps_f)

# Contaminant occurrences
sum(grepl("mitochondria", as.vector(tax_table(ps_f)[, "Family"]),
          ignore.case = TRUE), na.rm = TRUE) # 49 matches
sum(grepl("chloroplast", as.vector(tax_table(ps_f)[, "Order"]),
          ignore.case = TRUE), na.rm = TRUE) # 57 matches
sum(grepl("eukaryota", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 71 matches
sum(grepl("Unassigned", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 120

# Filtering out contaminants
ps_f <- subset_taxa(
  ps_f,
  !grepl("mitochondria", Family, ignore.case = TRUE) &
    !grepl("chloroplast", Order, ignore.case = TRUE) &
    !grepl("eukaryota", Domain, ignore.case = TRUE)
)
ps_f <- prune_taxa(taxa_sums(ps_f) > 0, ps_f)

# Checking filtered object
nsamples(ps_f) # 957
ntaxa(ps_f) # 11762
summary(sample_sums(ps_f)) # Min depth - 41051
```

## Rarefaction

### Rarefaction Curves

Different methods of "normalization" can be used before conducting statistical
analyses, rarefaction remains one of the best ones (Schloss, 2024). Since the
metadata required manipulation, rarefaction curves are produced in R (rather
than QIIME2).

```{r rarefaction_curves}
# Help from: https://www.youtube.com/watch?v=_OEdFjc1D9I
# Debugging with ELM (GPT 5.2), https://elm.edina.ac.uk/elm/elm
# OTU table as samples x taxa
otu_mat <- as(otu_table(ps_f), "matrix")
if (taxa_are_rows(ps_f)) otu_mat <- t(otu_mat)

# Batch vector
batch <- as.character(sample_data(ps_f)$batch)
names(batch) <- sample_names(ps_f)

# Align batch to OTU rows
batch <- batch[rownames(otu_mat)]

# Depth grid
step_size <- 1000
depths <- rowSums(otu_mat)

# Verifying 80k is within range
xmax <- max(c(depths, 80000), na.rm = TRUE)
depth_grid <- seq(step_size, as.integer(xmax), by = step_size)

# Compute expected richness at each depth
rich_mat <- suppressWarnings(
  sapply(depth_grid, function(d) vegan::rarefy(otu_mat, sample = d))
)

# Verifying steps as columns and samples as rows
colnames(rich_mat) <- as.character(depth_grid)
rownames(rich_mat) <- rownames(otu_mat)

# Long data frame for ggplot
rc_long <- as.data.frame(rich_mat, check.names = FALSE)
rc_long$sample_id <- rownames(rc_long)
rc_long <- rc_long %>%
  pivot_longer(-sample_id, names_to = "depth", values_to = "observed") %>%
  mutate(
    depth = as.numeric(depth),
    batch = batch[sample_id],
    sample_depth = depths[sample_id]
  )

# Stopping the plot when exceeding threshold (NA instead)
rc_long <- rc_long %>%
  mutate(
    sample_depth = depths[sample_id],
    observed = ifelse(depth > sample_depth, NA_real_, observed)
  )

# Defining panels
batches <- sort(unique(rc_long$batch))

# Global minimum
global_min <- min(rc_long$sample_depth, na.rm = TRUE)

threshold_df <- bind_rows(
  # Minimum within each batch
  rc_long %>%
    distinct(sample_id, batch, sample_depth) %>%
    group_by(batch) %>%
    summarize(threshold = min(sample_depth, na.rm = TRUE), .groups = "drop") %>%
    mutate(threshold_type = "min_in_batch"),

  # Global min replicated for each batch
  tibble(batch = batches,
         threshold_type = "global_min",
         threshold = global_min),

  # Fixed thresholds replicated for each batch
  tibble(batch = batches, threshold_type = "40000", threshold = 40000),
  tibble(batch = batches, threshold_type = "60000", threshold = 60000),
  tibble(batch = batches, threshold_type = "80000", threshold = 80000)
) %>%
  mutate(
    threshold_type = factor(
      threshold_type,
      levels = c("min_in_batch", "global_min",
                 "40000", "60000", "80000"),
      labels = c("Minimum Within Batch", "Global Minimum",
                 "40,000", "60,000", "80,000")
    )
  )

## Plots
rc_long$batch <- factor(rc_long$batch, levels = batches)
threshold_df$batch <- factor(threshold_df$batch, levels = batches)

# Plotting a figure for each batch
for (b in levels(rc_long$batch)) {

  th_b <- threshold_df %>%
    filter(batch == b) %>%
    select(threshold_type, threshold)

  df_b <- rc_long %>%
    filter(batch == b) %>%
    select(sample_id, depth, observed, sample_depth) %>%
    tidyr::crossing(th_b) %>%
    filter(depth <= threshold, depth <= sample_depth)

  p <- ggplot(df_b, aes(x = depth, y = observed, group = sample_id)) +
    geom_line(alpha = 0.3, linewidth = 0.45, color = "black", na.rm = TRUE) +
    geom_vline(aes(xintercept = threshold), linetype = 2, color = "red") +
    facet_wrap(~ threshold_type, nrow = 1, scales = "free_x") +
    theme_bw() +
    labs(
      title = paste0("Rarefaction Curves - Batch ", b),
      x = "Reads",
      y = "Expected observed ASVs"
    ) +
    theme(
      strip.text = element_text(size = 9),
      plot.title = element_text(face = "bold", hjust = 0.5)
    )

  ggsave(
    filename = file.path(paste0("rarefaction_plots_set2/rarefaction_batch_",
                                b, ".png")),
    plot = p,
    width = 14,
    height = 4.5,
    dpi = 300
  )
}

## Statistical summary - minimums
min_depth_by_batch <- data.frame(sample_data(ps_f),
                                 stringsAsFactors = FALSE) %>%
  rownames_to_column("sample_id") %>%
  mutate(
    depth = sample_sums(ps_f),
    batch = as.character(batch)
  ) %>%
  group_by(batch) %>%
  summarize(
    min_reads = min(depth),
    .groups = "drop"
  )
```

### Rarefying

As an initial exploration of batch effect after the chosen normalization method 
(rarefaction), a single subsampling was performed. Not used in the final
analysis.

```{r rarefying, eval=FALSE}
# Pruning phyloseq object
ps_f <- prune_samples(sample_sums(ps_f) >= global_min, ps_f)

# Subsampling
ps_r <- rarefy_even_depth(ps_f,
                          sample.size = global_min,
                          rngseed = 122863,
                          replace = FALSE,
                          verbose = FALSE)

# Sanity check (should be equal depth)
summary(sample_sums(ps_r))
```

## Batch Effect Post Rarefaction

Initial exploration of batch effect after accounting for uneven sampling depth
(rarefying). The same diversity metrics were calculated, with simplified
statistical tests to explore the effect. Not included in the final analysis.

```{r batch_post_rarefying, eval=FALSE}
# Add reads and alpha diversity metrics into sample_data
sample_data(ps_r)$depth <- sample_sums(ps_r)
alpha_post <- estimate_richness(ps_r,
                                measures = c("Observed",
                                             "Shannon",
                                             "InvSimpson"))
sample_data(ps_r)$observed <- alpha_post[sample_names(ps_r), "Observed"]
sample_data(ps_r)$shannon  <- alpha_post[sample_names(ps_r), "Shannon"]
sample_data(ps_r)$simpson  <- alpha_post[sample_names(ps_r), "InvSimpson"]

# Faith's PD
otu_post <- as.matrix(otu_table(ps_r))
if (taxa_are_rows(ps_r)) otu_post <- t(otu_post)
pd_post <- picante::pd(otu_post, phy_tree(ps_r))
faith_post <- pd_post$PD
names(faith_post) <- rownames(pd_post)
sample_data(ps_r)$faith_pd <- faith_post[sample_names(ps_r)]

# Single plotting table
meta_batch_post <- data.frame(sample_data(ps_r), stringsAsFactors = FALSE) %>%
  rownames_to_column("sample_id") %>%
  filter(!is.na(batch)) %>%
  mutate(batch = factor(batch))

# Reads by batch
batch_reads_post <- ggplot(meta_batch_post,
                           aes(x = batch, y = depth, fill = batch)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "GnBu") +
  stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  ggtitle("Number of Reads by Sequencing Batch") +
  ylab("Number of Reads") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "bottom"
  )

# Summary statistics by batch
reads_summary_batch_post <- meta_batch_post %>%
  group_by(batch) %>%
  summarize(
    n = n(),
    min = min(depth, na.rm = TRUE),
    max = max(depth, na.rm = TRUE),
    median = median(depth, na.rm = TRUE),
    mean = mean(depth, na.rm = TRUE),
    .groups = "drop"
  )

# Alpha diversity boxplots by batch
batch_alpha_post <- ggarrange(
  # Observed ASVs
  ggplot(meta_batch_post, aes(x = batch, y = observed, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Shannon's Index
  ggplot(meta_batch_post, aes(x = batch, y = shannon, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(meta_batch_post, aes(x = batch, y = simpson, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(meta_batch_post, aes(x = batch, y = faith_pd, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 3, nrow = 1,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
batch_alpha_post <- annotate_figure(batch_alpha_post,
                                    top = text_grob(paste0("Alpha Diversity ",
                                                           "Metrics per Batch ",
                                                           "Post Rarefaction"),
                                                    face = "bold", size = 14))

# Alpha diversity summary statistics by batch
alpha_summary_batch_post <- meta_batch_post %>%
  group_by(batch) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(simpson, na.rm = TRUE),
    max_simpson = max(simpson, na.rm = TRUE),
    median_simpson = median(simpson, na.rm = TRUE),
    mean_simpson = mean(simpson, na.rm = TRUE),
    min_faith = min(faith_pd, na.rm = TRUE),
    max_faith = max(faith_pd, na.rm = TRUE),
    median_faith = median(faith_pd, na.rm = TRUE),
    mean_faith = mean(faith_pd, na.rm = TRUE),
    .groups = "drop"
  )

# Beta diversity and PCoA by batch
dist_bray_post <- phyloseq::distance(ps_r, method = "bray")
dist_jac_post <- phyloseq::distance(ps_r, method = "jaccard", binary = TRUE)

# Previous pruning could have led to a non-dichotomic tree
# https://github.com/joey711/phyloseq/issues/936
phy_tree(ps_r) <- ape::multi2di(phy_tree(ps_r))
dist_uwu_post <- phyloseq::UniFrac(ps_r, weighted = FALSE, normalized = TRUE)
dist_wu_post <- phyloseq::UniFrac(ps_r, weighted = TRUE,  normalized = TRUE)

# Ordinations
set.seed(122863)
ord_bray_post <- ordinate(ps_r, method = "PCoA", distance = dist_bray_post)
set.seed(122863)
ord_jac_post <- ordinate(ps_r, method = "PCoA", distance = dist_jac_post)
set.seed(122863)
ord_uwu_post <- ordinate(ps_r, method = "PCoA", distance = dist_uwu_post)
set.seed(122863)
ord_wu_post <- ordinate(ps_r, method = "PCoA", distance = dist_wu_post)

batch_beta_post <- ggarrange(
  plot_ordination(ps_r, ord_bray_post, color = "batch") +
    geom_point(size = 2) +
    labs(color = "Sequencing batch", caption = "Bray-Curtis PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  plot_ordination(ps_r, ord_jac_post, color = "batch") +
    geom_point(size = 2) +
    labs(color = "Sequencing batch", caption = "Jaccard (binary) PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  plot_ordination(ps_r, ord_uwu_post, color = "batch") +
    geom_point(size = 2) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Sequencing batch", caption = "Unweighted UniFrac PCoA"),

  plot_ordination(ps_r, ord_wu_post, color = "batch") +
    geom_point(size = 2) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Sequencing batch", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
batch_beta_post <- annotate_figure(batch_beta_post,
                                   top = text_grob(paste0("Beta Diversity ",
                                                          "Metrics per Batch ",
                                                          "Post Rarefaction"),
                                                   face = "bold", size = 14))

## Exploratory statistical tests
# For beta metrics only
# Exploration only - tests do not account for covariates
sd <- data.frame(sample_data(ps_r))

# PERMANOVA
set.seed(122863)
sd <- sd[labels(dist_bray_post), , drop = FALSE]
permanova_batch_bray <- adonis2(dist_bray_post ~ batch,
                                data = sd,
                                permutations = 9999)
set.seed(122863)
sd <- sd[labels(dist_jac_post), , drop = FALSE]
permanova_batch_jaccard <- adonis2(dist_jac_post ~ batch,
                                   data = sd,
                                   permutations = 9999)
set.seed(122863)
sd <- sd[labels(dist_wu_post), , drop = FALSE]
permanova_batch_wu <- adonis2(dist_wu_post ~ batch,
                              data = sd,
                              permutations = 9999)
set.seed(122863)
sd <- sd[labels(dist_uwu_post), , drop = FALSE]
permanova_batch_uwu <- adonis2(dist_uwu_post ~ batch,
                               data = sd,
                               permutations = 9999)

# Dispersion (PERDISP)
# Helper function to run dispersion
# Inspired by: https://www.youtube.com/watch?v=oLf0EpMJ4yA
run_dispersion <- function(dist, group, permutations = 9999) {

  bd <- betadisper(dist, group = group)
  pt <- permutest(bd, permutations = permutations, pairwise = TRUE)

  list(betadisper = bd, permutest = pt)
}

# PERMDISP tests for significant PERMANOVA
sd <- sd[labels(dist_jac_post), , drop = FALSE]
set.seed(122863)
disp_jac  <- run_dispersion(dist_jac_post, sd$batch)

sd <- sd[labels(dist_uwu_post), , drop = FALSE]
set.seed(122863)
disp_uwu  <- run_dispersion(dist_uwu_post, sd$batch)
```

## Final Descriptive Statistics

Final descriptive information for samples that will proceed into the statistical
analysis after filtering. Counts and percentages are calculated per categorical
variable. Continuous age statistics were also calculated.

```{r descriptive_stats}
md_f <- sample_data(ps_f)
md_f <- as(md_f, "data.frame")

# Sex --------------------------------------------------------------------------
sex_stats <- md_f %>%
  group_by(sex) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Age --------------------------------------------------------------------------
# Categorical
age_cat <- md_f %>%
  group_by(age_categorical) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Continuous
# TODO: perhaps plot distribution
age_cont <- md_f %>%
  summarize(
    Minimum = min(age_years),
    Maximum = max(age_years),
    Median = median(age_years),
    Mean = mean(age_years),
    SD = sd(age_years)
  )

# Batch ------------------------------------------------------------------------
# TODO: perhaps look at variables across batches
batch_stats <- md_f %>%
  group_by(batch) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Antibiotics ------------------------------------------------------------------
antibiotics_stats <- md_f %>%
  group_by(antibiotics) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Neutering  Status ------------------------------------------------------------
neutering_stats <- md_f %>%
  group_by(neutered) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Combined w/ sex
# Twice as many intact males than females
neutering_sex_stats <- md_f %>%
  count(neutered, sex, name = "n") %>%
  mutate(perc = 100 * n / sum(n))

# Physical Activity ------------------------------------------------------------
active_stats <- md_f %>%
  group_by(active) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Body Condition ---------------------------------------------------------------
body_condition_stats <- md_f %>%
  group_by(body_condition) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Diet -------------------------------------------------------------------------
diet_stats <- md_f %>%
  group_by(diet) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Breed ------------------------------------------------------------------------
breed_stats <- md_f %>%
  group_by(breed_clean) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Biotics ----------------------------------------------------------------------
biotics_stats <- md_f %>%
  group_by(biotics) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Healthy / Symptoms -----------------------------------------------------------
# Categories are not mutually exclusive
symptoms_stats <- md_f %>%
  summarize(
    healthy = sum(healthy %in% "TRUE"),
    gastrointestinal = sum(gastro %in% "TRUE"),
    skin = sum(skin %in% "TRUE"),
    neurological_behavioral = sum(neuro %in% "TRUE")
  )

# Vaccination ------------------------------------------------------------------
vaccination_stats <- md_f %>%
  group_by(vaccination) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Worming ----------------------------------------------------------------------
worming_stats <- md_f %>%
  group_by(worming) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Location ---------------------------------------------------------------------
location_stats <- md_f %>%
  group_by(location) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)
```

## Alpha Diversity

First assessment of host factors associations with the microbiome is performed
with alpha diversity metrics (within-sample diversity).

```{r avg_alpha}
# Calculating alpha diversity metrics with repeated subsampling
# From:
# https://bioconductor.posit.co/packages/3.23/bioc/manuals/mia/man/mia.pdf

if (file.exists("avg_results_set2/avg_alpha.RDS")) {
  tse <- readRDS("avg_results_set2/avg_alpha.RDS")
} else {
  # Converting to mia-compatible object
  tse <- convertFromPhyloseq(ps_f)

  # 100 iterations, requesting five metrics
  # Chao1 was initially considered but not included in the final analysis
  # because of:
  # https://pubmed.ncbi.nlm.nih.gov/38869966/
  set.seed(122863)
  tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = c("observed", "chao1", "shannon", "inverse_simpson", "faith"),
    sample = global_min,
    niter = 100
  )

  # Checking and saving object
  head(colData((tse)))
  saveRDS(object = tse,
          file = "avg_results_set2/avg_alpha.RDS")
}

# Metadata Object
md_avg <- as.data.frame(colData(tse))
```

### Observed ASVs

```{r observed}
# Distribution -----------------------------------------------------------------
observed_dist <- ggplot(md_avg, aes(x = observed)) +
  geom_histogram() +
  ggtitle("Richness Distribution") +
  xlab("Observed ASVs") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
# Does not look very normal with right tail

# Descriptive statistics -------------------------------------------------------
observed_stats <- md_avg %>%
  summarize(
    Minimum = min(observed),
    Maximum = max(observed),
    Median = median(observed),
    Mean = mean(observed),
    SD = sd(observed)
  )

# Linear model -----------------------------------------------------------------
observed_lm <- lm(observed ~ sex + breed_clean + neutered +
                    active + body_condition + diet + biotics +
                    gastro + skin + neuro + healthy + worming +
                    antibiotics + location + age_categorical + batch,
                  data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(observed_lm, which = c(1, 2, 4))

# Log-transformation -----------------------------------------------------------
md_avg$observed_log <- log(md_avg$observed)
observed_log <- lm(observed_log ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(observed_log, which = c(1, 2, 4))

# Negative binomial (massive overdispersion with poisson) ----------------------
observed_nb <- glm.nb(observed ~ sex + breed_clean + neutered +
                        active + body_condition + diet + biotics +
                        gastro + skin + neuro + healthy + worming +
                        antibiotics + location + age_categorical + batch,
                      data = md_avg)

# Checking model fit
summary(observed_nb)
testDispersion(observed_nb)
# Significant dispersion
# Can occur with with large sample size according to:
# https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html

simulated_observed_nb <- simulateResiduals(fittedModel = observed_nb,
                                           plot = FALSE)
plot(simulated_observed_nb)

# Individual parameters --------------------------------------------------------
plotResiduals(simulationOutput, form = ~ batch)
plotResiduals(simulationOutput, form = ~ breed_clean)
plotResiduals(simulationOutput, form = ~ sex)
plotResiduals(simulationOutput, form = ~ neutered)
plotResiduals(simulationOutput, form = ~ active)
plotResiduals(simulationOutput, form = ~ body_condition)
plotResiduals(simulationOutput, form = ~ diet)
plotResiduals(simulationOutput, form = ~ biotics)
plotResiduals(simulationOutput, form = ~ antibiotics)
plotResiduals(simulationOutput, form = ~ worming)
plotResiduals(simulationOutput, form = ~ healthy)
plotResiduals(simulationOutput, form = ~ location)
plotResiduals(simulationOutput,
              form = ~ age_categorical)
plotResiduals(simulationOutput, form = ~ gastro)
plotResiduals(simulationOutput, form = ~ skin)
plotResiduals(simulationOutput, form = ~ neuro)
```

### Chao1

```{r chao1, echo=FALSE}
# Excluded because of:
# https://pubmed.ncbi.nlm.nih.gov/38869966/
# Distribution -----------------------------------------------------------------
chao1_dist <- ggplot(md_avg, aes(x = chao1)) +
  geom_histogram() +
  ggtitle("Chao1 Distribution") +
  xlab("Chao1") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
chao1_stats <- md_avg %>%
  summarize(
    Minimum = min(chao1),
    Maximum = max(chao1),
    Median = median(chao1),
    Mean = mean(chao1),
    SD = sd(chao1)
  )
# Does not look very normal with right tail

# Linear model -----------------------------------------------------------------
chao1_lm <- lm(chao1 ~ sex + breed_clean + neutered +
                 active + body_condition + diet + biotics +
                 gastro + skin + neuro + healthy + worming +
                 antibiotics + location + age_categorical + batch,
               data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(chao1_lm, which = c(1, 2, 4))

# lmer -------------------------------------------------------------------------
chao1_lmer <- lmerTest::lmer(chao1 ~ sex + breed_clean + neutered +
                               active + body_condition + diet + biotics +
                               gastro + skin + neuro + healthy + worming +
                               antibiotics + location + age_categorical +
                               (1 | batch),
                             data = md_avg)

# Checking model fit
testDispersion(chao1_lmer)
simulated_chao_lmer <- simulateResiduals(fittedModel = chao1_lmer,
                                         plot = FALSE)
plot(simulated_chao_lmer)

# Log-transformation -----------------------------------------------------------
md_avg$chao1_log <- log(md_avg$chao1)
chao1_log <- lm(chao1_log ~ sex + breed_clean + neutered +
                  active + body_condition + diet + biotics +
                  gastro + skin + neuro + healthy + worming +
                  antibiotics + location + age_categorical + batch,
                data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(chao1_log, which = c(1, 2, 4))

# Gamma GLM --------------------------------------------------------------------
chao1_gamma <- glm(chao1 ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   family = Gamma(link = "log"), data = md_avg)

# Checking model fit
testDispersion(chao1_gamma)
simulated_chao_gamma <- simulateResiduals(fittedModel = chao1_gamma,
                                          plot = FALSE)
plot(simulated_chao_gamma)

# Gamma GLMER ------------------------------------------------------------------
chao1_glmer <- glmer(chao1 ~ sex + breed_clean + neutered +
                       active + body_condition + diet + biotics +
                       gastro + skin + neuro + healthy + worming +
                       antibiotics + location + age_categorical + (1 | batch),
                     family = Gamma(link = "log"), data = md_avg)

# Checking model fit
testDispersion(chao1_glmer)
simulated_chao_glmer <- simulateResiduals(fittedModel = chao1_glmer,
                                          plot = FALSE)
plot(simulated_chao_glmer)
```

### Shannon's Index

```{r shannon}
# Distribution -----------------------------------------------------------------
shannon_distribution <- ggplot(md_avg, aes(x = shannon)) +
  geom_histogram() +
  ggtitle("Shannon Diversity Distribution") +
  xlab("Shannon's Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
shannon_stats <- md_avg %>%
  summarize(
    Minimum = min(shannon),
    Maximum = max(shannon),
    Median = median(shannon),
    Mean = mean(shannon),
    SD = sd(shannon)
  )

# Linear model -----------------------------------------------------------------
shannon_lm <- lm(shannon ~ sex + breed_clean + neutered +
                   active + body_condition + diet + biotics +
                   gastro + skin + neuro + healthy + worming +
                   antibiotics + location + age_categorical + batch,
                 data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(shannon_lm, which = c(1, 2, 4))

# lmer -------------------------------------------------------------------------
# Boundary is singular
shannon_lmer <- lmerTest::lmer(shannon ~ sex + breed_clean + neutered +
                                 active + body_condition + diet + biotics +
                                 gastro + skin + neuro + healthy + worming +
                                 antibiotics + location + age_categorical +
                                 (1 | batch),
                               data = md_avg)
```

### Simpson's Index

```{r inverse_simpson}
# Distribution -----------------------------------------------------------------
simpson_distribution <- ggplot(md_avg, aes(x = inverse_simpson)) +
  geom_histogram() +
  ggtitle("Inverse Simpson Diversity Distribution") +
  xlab("Inverse Simpson Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
# Does not look very normal with left tail
# But not too bad, a weaker transformation (e.g., square-root) might be better

# Descriptive statistics -------------------------------------------------------
simpson_stats <- md_avg %>%
  summarize(
    Minimum = min(inverse_simpson),
    Maximum = max(inverse_simpson),
    Median = median(inverse_simpson),
    Mean = mean(inverse_simpson),
    SD = sd(inverse_simpson)
  )

# Linear model -----------------------------------------------------------------
simpson_lm <- lm(inverse_simpson ~ sex + breed_clean + neutered +
                   active + body_condition + diet + biotics +
                   gastro + skin + neuro + healthy + worming +
                   antibiotics + location + age_categorical + batch,
                 data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(simpson_lm, which = c(1, 2, 4))

# Log Transformation -----------------------------------------------------------
md_avg$simpson_log <- log(md_avg$inverse_simpson)
simpson_log <- lm(simpson_log ~ sex + breed_clean + neutered +
                    active + body_condition + diet + biotics +
                    gastro + skin + neuro + healthy + worming +
                    antibiotics + location + age_categorical + batch,
                  data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(simpson_log, which = c(1, 2, 4))

# Square-root Transformation ---------------------------------------------------
md_avg$simpson_sqrt <- sqrt(md_avg$inverse_simpson)
simpson_sqrt <- lm(simpson_sqrt ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(simpson_sqrt, which = c(1, 2, 4))

# Transformed distribution -----------------------------------------------------
simpson_sqrt_dist <- ggplot(md_avg, aes(x = simpson_sqrt)) +
  geom_histogram() +
  ggtitle("Square-root Inverse Simpson Distribution") +
  xlab("Square-root Inverse Simpson Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# lmer -------------------------------------------------------------------------
# Boundary is singular
simpson_lmer <- lmerTest::lmer(simpson_sqrt ~ sex + breed_clean + neutered +
                                 active + body_condition + diet + biotics +
                                 gastro + skin + neuro + healthy + worming +
                                 antibiotics + location + age_categorical +
                                 (1 | batch),
                               data = md_avg)
```

### Faith's PD

```{r faith_pd}
# Distribution -----------------------------------------------------------------
faith_distribution <- ggplot(md_avg, aes(x = faith)) +
  geom_histogram() +
  ggtitle("Faith's Phylogenetic Distance Distribution") +
  xlab("Faith's PD") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
# Does not look very normal with right tail

# Descriptive statistics -------------------------------------------------------
faith_stats <- md_avg %>%
  summarize(
    Minimum = min(faith),
    Maximum = max(faith),
    Median = median(faith),
    Mean = mean(faith),
    SD = sd(faith)
  )

# Linear model -----------------------------------------------------------------
faith_lm <- lm(faith ~ sex + breed_clean + neutered +
                 active + body_condition + diet + biotics +
                 gastro + skin + neuro + healthy + worming +
                 antibiotics + location + age_categorical + batch,
               data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(faith_lm, which = c(1, 2, 4))

# Log Transformation -----------------------------------------------------------
md_avg$faith_log <- log(md_avg$faith)
faith_log <- lm(faith_log ~ sex + breed_clean + neutered +
                  active + body_condition + diet + biotics +
                  gastro + skin + neuro + healthy + worming +
                  antibiotics + location + age_categorical + batch,
                data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(faith_log, which = 1)
plot(faith_log, which = 2)
# Adding a 1:1 line for the Q-Q plot
abline(0, 1)
plot(faith_log, which = 4)

# Transformed distribution -----------------------------------------------------
faith_log_dist <- ggplot(md_avg, aes(x = faith_log)) +
  geom_histogram() +
  ggtitle("Log-transformed Faith PD Distribution") +
  xlab("Log-transformed Faith PD") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# lmer -------------------------------------------------------------------------
faith_lmer <- lmerTest::lmer(faith_log ~ sex + breed_clean + neutered +
                               active + body_condition + diet + biotics +
                               gastro + skin + neuro + healthy + worming +
                               antibiotics + location + age_categorical +
                               (1 | batch),
                             data = md_avg)

# Check model fit
testDispersion(faith_lmer)
simulated_faith_lmer <- simulateResiduals(fittedModel = faith_lmer,
                                          plot = FALSE)
plot(simulated_faith_lmer)
```

### Model Summary

Summary statistics across models. 

```{r alpha_models}
# Anova type II for models w/o interactions
# From:
# https://www.r-bloggers.com/2011/03/anova-%E2%80%93-type-iiiiii-ss-explained/

# Helper function for post-hoc comparisons for significant variables -----------
# Help from:
# https://stats.stackexchange.com/questions/102857/post-hoc-after-glm
# https://cran.r-project.org/web/packages/emmeans/vignettes/AQuickStart.html
emm_posthoc <- function(fit,
                        factors,
                        adjust = "BH") {
  mf <- model.frame(fit)

  # Factor variables in the model
  factor_vars <- names(mf)[vapply(mf, is.factor, logical(1))]

  out <- lapply(factors, function(term) {

    # Simplifying calculation grid
    # Keeping other covariates averaged across levels, proportionally-weighted
    nuisance_vars <- setdiff(factor_vars, term)
    emm <- emmeans(
      fit,
      as.formula(paste("~", term)),
      nuisance = nuisance_vars,
      weights = "proportional",
      type = "response"
    )

    # Pairwise comparisons
    prs <- pairs(emm, adjust = adjust, type = "response")

    df <- as.data.frame(summary(prs, infer = TRUE))
    df$term <- term
    df
  })

  do.call(rbind, out)
}

# Observed - negative binomial -------------------------------------------------
car::Anova(observed_nb,
           test.statistic = "LR")

# Post-hoc for significant factors
# Breed and batch were also significant, but may inflate the results
sig_observed <- c("active", "body_condition",
                  "location", "diet", "age_categorical",
                  "skin", "healthy")
post_observed <- emm_posthoc(fit = observed_nb,
                             factors = sig_observed,
                             adjust = "BH")
saveRDS(object = post_observed, file = "alpha_stats_set2/post_observed.RDS")

# Chao1 - gamma ----------------------------------------------------------------
# Excluded from final analysis
car::Anova(chao1_gamma,
           test.statistic = "LR")

# Post-hoc for significant factors
# Breed and batch were also significant, but may inflate the results
sig_chao <- c("diet", "age_categorical", "body_condition",
              "skin", "healthy")
post_chao <- emm_posthoc(fit = chao1_gamma,
                         factors = sig_chao,
                         adjust = "BH")
saveRDS(object = post_chao, file = "alpha_stats_set2/post_chao.RDS")


# Shannon - linear model -------------------------------------------------------
car::Anova(shannon_lm)

# Post-hoc for significant factors
# Breed was also significant, but may inflate the results
sig_shannon <- c("antibiotics", "age_categorical", "diet",
                 "sex", "skin")
post_shannon <- emm_posthoc(fit = shannon_lm,
                            factors = sig_shannon,
                            adjust = "BH")
saveRDS(object = post_shannon, file = "alpha_stats_set2/post_shannon.RDS")

# Inverse Simpson - sqrt transformation ----------------------------------------
car::Anova(simpson_sqrt)

# Post-hoc for significant factors
# Breed was also significant, but may inflate the results
sig_simpson <- c("antibiotics", "age_categorical",
                 "sex")
post_simpson <- emm_posthoc(fit = simpson_sqrt,
                            factors = sig_simpson,
                            adjust = "BH")
saveRDS(object = post_simpson, file = "alpha_stats_set2/post_simpson.RDS")

# Faith's PD - log transformation ----------------------------------------------
car::Anova(faith_log)

# Post-hoc for significant factors
# Breed and age were also significant, but may inflate the results
sig_faith <- c("active", "diet", "age_categorical")
post_faith <- emm_posthoc(fit = faith_log,
                          factors = sig_faith,
                          adjust = "BH")
saveRDS(object = post_faith, file = "alpha_stats_set2/post_faith.RDS")
```

### Plots per Variable

Plots of all untransformed alpha metrics for significant covariates.

#### Sex

```{r alpha_sex}
# Plots
alpha_sex <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = sex, y = observed, fill = sex)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Chao1
  # ggplot(md_avg, aes(x = sex, y = chao1, fill = sex)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = sex, y = shannon, fill = sex)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = sex, y = inverse_simpson, fill = sex)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = sex, y = faith, fill = sex)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_sex <- annotate_figure(alpha_sex,
                             top = text_grob(paste0("Sex - Alpha ",
                                                    "Diversity Metrics"),
                                             face = "bold", size = 14))

# Stats
alpha_summary_sex <- md_avg %>%
  group_by(sex) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Breed

```{r alpha_breed}
# TODO: think about plotting 11 categories...
```

#### Physical Activity

```{r alpha_active}
# Plots
alpha_active <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = active, y = observed, fill = active)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = active, y = chao1, fill = active)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = active, y = shannon, fill = active)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = active, y = inverse_simpson, fill = active)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = active, y = faith, fill = active)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_active <- annotate_figure(alpha_active,
                                top = text_grob(paste0("Physical Activity ",
                                                       "- Alpha Diversity ",
                                                       "Metrics"),
                                                face = "bold", size = 14))

# Stats
alpha_summary_active <- md_avg %>%
  group_by(active) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Body Condition

```{r alpha_body_condition}
## Observed ASVs only...
# Plots
alpha_body_condition <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = body_condition, y = observed, fill = body_condition)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = body_condition, y = chao1, fill = body_condition)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = body_condition, y = shannon, fill = body_condition)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = body_condition,
                     y = inverse_simpson,
                     fill = body_condition)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = body_condition, y = faith, fill = body_condition)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_body_condition <- annotate_figure(alpha_body_condition,
                                        top = text_grob(paste0("Body Condition",
                                                               " - Alpha ",
                                                               "Diversity ",
                                                               "Metrics"),
                                                        face = "bold",
                                                        size = 14))

# Stats
alpha_summary_body_condition <- md_avg %>%
  group_by(body_condition) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Diet

```{r alpha_diet}
# Plots
alpha_diet <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = diet, y = observed, fill = diet)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = diet, y = chao1, fill = diet)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = diet, y = shannon, fill = diet)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = diet, y = inverse_simpson, fill = diet)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = diet, y = faith, fill = diet)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_diet <- annotate_figure(alpha_diet,
                              top = text_grob(paste0("Diet - ",
                                                     "Alpha Diversity ",
                                                     "Metrics"),
                                              face = "bold", size = 14))

# Stats
alpha_summary_active <- md_avg %>%
  group_by(diet) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Healthy

```{r alpha_healthy}
## Only in richness metrics
# Plots
alpha_healthy <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = healthy, y = observed, fill = healthy)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = healthy, y = chao1, fill = healthy)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = healthy, y = shannon, fill = healthy)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = healthy, y = inverse_simpson, fill = healthy)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = healthy, y = faith, fill = healthy)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_healthy <- annotate_figure(alpha_healthy,
                                 top = text_grob(paste0("Health - ",
                                                        "Alpha Diversity ",
                                                        "Metrics"),
                                                 face = "bold", size = 14))

# Stats
alpha_summary_healthy <- md_avg %>%
  group_by(healthy) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Skin-related Symptoms

```{r alpha_skin}
# Plots
alpha_skin <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = skin, y = observed, fill = skin)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = skin, y = chao1, fill = skin)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = skin, y = shannon, fill = skin)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = skin, y = inverse_simpson, fill = skin)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = skin, y = faith, fill = skin)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_skin <- annotate_figure(alpha_skin,
                              top = text_grob(paste0("Skin-related Symptoms ",
                                                     "- Alpha Diversity ",
                                                     "Metrics"),
                                              face = "bold", size = 14))

# Stats
alpha_summary_skin <- md_avg %>%
  group_by(skin) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Living Environment

```{r alpha_location}
## Richness metrics only...
# Plots
alpha_location <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = location, y = observed, fill = location)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = location, y = chao1, fill = location)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = location, y = shannon, fill = location)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = location, y = inverse_simpson, fill = location)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = location, y = faith, fill = location)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_location <- annotate_figure(alpha_location,
                                  top = text_grob(paste0("Living Environment ",
                                                         "- Alpha Diversity ",
                                                         "Metrics"),
                                                  face = "bold", size = 14))

# Stats
alpha_summary_location <- md_avg %>%
  group_by(location) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Antibiotics

```{r alpha_antibiotics}
# Plots
alpha_antibiotics <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = antibiotics, y = observed, fill = antibiotics)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = antibiotics, y = chao1, fill = antibiotics)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = antibiotics, y = shannon, fill = antibiotics)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = antibiotics,
                     y = inverse_simpson,
                     fill = antibiotics)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = antibiotics, y = faith, fill = antibiotics)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_antibiotics <- annotate_figure(alpha_antibiotics,
                                     top = text_grob(paste0("Antibiotics - ",
                                                            "Alpha Diversity ",
                                                            "Metrics"),
                                                     face = "bold", size = 14))

# Stats
alpha_summary_antibiotics <- md_avg %>%
  group_by(antibiotics) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Age

```{r alpha_age}
# Plots
alpha_age <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = age_categorical,
                     y = observed,
                     fill = age_categorical)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = age_categorical,
  #                    y = chao1,
  #                    fill = age_categorical)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = age_categorical,
                     y = shannon,
                     fill = age_categorical)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = age_categorical,
                     y = inverse_simpson,
                     fill = age_categorical)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = age_categorical, y = faith, fill = age_categorical)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_age <- annotate_figure(alpha_age,
                             top = text_grob(paste0("Age - ",
                                                    "Alpha Diversity ",
                                                    "Metrics"),
                                             face = "bold", size = 14))

# Stats
alpha_summary_age <- md_avg %>%
  group_by(age_categorical) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

#### Batch

```{r alpha_batch}
# Plots
alpha_batch <- ggarrange(

  # Observed ASVs
  ggplot(md_avg, aes(x = batch, y = observed, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    #scale_fill_brewer(palette = "GnBu") +
    ylab("Observed ASVs") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # # Chao1
  # ggplot(md_avg, aes(x = batch, y = chao1, fill = batch)) +
  #   geom_boxplot() +
  #   stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
  #   #scale_fill_brewer(palette = "GnBu") +
  #   ylab("Chao1") +
  #   theme(
  #     axis.title.x = element_blank(),
  #     axis.text.x = element_blank(),
  #     axis.ticks.x = element_blank(),
  #     axis.title.y = element_text(face = "bold", size = 12),
  #     legend.position = "none"
  #   ),

  # Shannon's Index
  ggplot(md_avg, aes(x = batch, y = shannon, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    #scale_fill_brewer(palette = "GnBu") +
    ylab("Shannon's Index") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Inverse Simpson
  ggplot(md_avg, aes(x = batch, y = inverse_simpson, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    #scale_fill_brewer(palette = "GnBu") +
    ylab("Inverse Simpson") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  # Faith's PD
  ggplot(md_avg, aes(x = batch, y = faith, fill = batch)) +
    geom_boxplot() +
    stat_summary(fun = "mean", geom = "point", shape = 18, size = 3) +
    #scale_fill_brewer(palette = "GnBu") +
    ylab("Faith's PD") +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.y = element_text(face = "bold", size = 12),
      legend.position = "none"
    ),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
alpha_batch <- annotate_figure(alpha_batch,
                               top = text_grob(paste0("Batch - ",
                                                      "Alpha Diversity ",
                                                      "Metrics"),
                                               face = "bold", size = 14))

# Stats
alpha_summary_batch <- md_avg %>%
  group_by(batch) %>%
  summarize(
    n = n(),
    min_observed = min(observed, na.rm = TRUE),
    max_observed = max(observed, na.rm = TRUE),
    median_observed = median(observed, na.rm = TRUE),
    mean_observed = mean(observed, na.rm = TRUE),
    min_chao1 = min(chao1, na.rm = TRUE),
    max_chao1 = max(chao1, na.rm = TRUE),
    median_chao1 = median(chao1, na.rm = TRUE),
    mean_chao1 = mean(chao1, na.rm = TRUE),
    min_shannon = min(shannon, na.rm = TRUE),
    max_shannon = max(shannon, na.rm = TRUE),
    median_shannon = median(shannon, na.rm = TRUE),
    mean_shannon = mean(shannon, na.rm = TRUE),
    min_simpson = min(inverse_simpson, na.rm = TRUE),
    max_simpson = max(inverse_simpson, na.rm = TRUE),
    median_simpson = median(inverse_simpson, na.rm = TRUE),
    mean_simpson = mean(inverse_simpson, na.rm = TRUE),
    min_faith = min(faith, na.rm = TRUE),
    max_faith = max(faith, na.rm = TRUE),
    median_faith = median(faith, na.rm = TRUE),
    mean_faith = mean(faith, na.rm = TRUE),
    .groups = "drop"
  )
```

## Beta Diversity

Second assessment of host factors associations with the microbiome is performed
with beta diversity metrics (between-sample diversity). PERMANOVA tests are
performed, and dispersion is estimated with PERMDISP.

```{r avg_beta}
# Getting averaged metrics and saving objects ----------------------------------
# From:
# https://bioconductor.posit.co/packages/3.23/bioc/manuals/mia/man/mia.pdf
# 100 iterations, four metrics

# Bray-Curtis
if (file.exists("avg_results_set2/avg_bray.RDS")) {
  avg_bray <- readRDS("avg_results_set2/avg_bray.RDS")
} else {
  set.seed(122863)
  avg_bray <- getDissimilarity(
    tse,
    method = "bray",
    assay.type = "counts",
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_bray,
          file = "avg_results_set2/avg_bray.RDS")
}

# Jaccard
if (file.exists("avg_results_set2/avg_jac.RDS")) {
  avg_jac <- readRDS("avg_results_set2/avg_jac.RDS")
} else {
  set.seed(122863)
  avg_jac <- getDissimilarity(
    tse,
    method = "jaccard",
    assay.type = "counts",
    binary = TRUE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_jac,
          file = "avg_results_set2/avg_jac.RDS")
}

# Unweighted UniFrac
if (file.exists("avg_results_set2/avg_uwu.RDS")) {
  avg_uwu <- readRDS("avg_results_set2/avg_uwu.RDS")
} else {
  set.seed(122863)
  avg_uwu <- getDissimilarity(
    tse, method = "unifrac",
    name = "uwu",
    assay.type = "counts",
    weighted = FALSE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_uwu,
          file = "avg_results_set2/avg_uwu.RDS")
}

# Weighted UniFrac
if (file.exists("avg_results_set2/avg_wu.RDS")) {
  avg_wu <- readRDS("avg_results_set2/avg_wu.RDS")
} else {
  set.seed(122863)
  avg_wu <- getDissimilarity(
    tse, method = "unifrac",
    name = "wu",
    assay.type = "counts",
    weighted = TRUE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_wu,
          file = "avg_results_set2/avg_wu.RDS")
}

# Helper function to run dispersion --------------------------------------------
# Inspired by: https://www.youtube.com/watch?v=oLf0EpMJ4yA
run_dispersion <- function(dist, group, permutations = 999) {

  bd <- betadisper(dist, group = group)
  pt <- permutest(bd, permutations = permutations)

  list(betadisper = bd, permutest = pt, p_global = pt$tab["Groups", "Pr(>F)"])
}

# Helper function for pairwise PERMANOVA comparisons ---------------------------

# Adapted from:
# https://gist.github.com/mcgoodman/58c9d1257fd1625954a4ffa1c3301939
# And: the pairwise.adonis2() function
# Using ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
pairwise_permanova <- function(dist, factor_name, covars = NULL, meta,
                               permutations = 999, adjust = "BH",
                               seed = 122863) {

  # Aligning metadata rows to the distance object
  meta <- meta[labels(dist), , drop = FALSE]
  dmat <- as.matrix(dist)

  # Factor vector and all-level combinations
  fac <- droplevels(as.factor(meta[[factor_name]]))
  pairs <- combn(levels(fac), 2, simplify = FALSE)

  # Initializing a results list
  results <- list()

  # Iterating through the possible combinations, calculating the statistics
  # for each variable restricted to a pair of values
  for (pr in pairs) {
    # Subsetting to the two levels being compared
    keep <- fac %in% pr
    d_sub <- as.dist(dmat[keep, keep])
    m_sub <- droplevels(meta[keep, , drop = FALSE])

    # Keeping only covariates that vary within the pair (more than one level)
    good_covars <- covars[sapply(covars, function(cv) {
      nlevels(droplevels(factor(m_sub[[cv]]))) > 1
    })]

    # Building formula
    rhs <- paste(c(good_covars, factor_name), collapse = " + ")
    form <- as.formula(paste("d_sub ~", rhs))

    # Marginal test w/ adjustment for covariates
    set.seed(seed)
    fit <- adonis2(form, data = m_sub, by = "margin",
                   permutations = permutations)

    # Tidy results
    results[[paste(pr, collapse = "_vs_")]] <- data.frame(
      comparison = paste(pr, collapse = "_vs_"),
      n1 = sum(m_sub[[factor_name]] == pr[1]),
      n2 = sum(m_sub[[factor_name]] == pr[2]),
      R2 = fit[factor_name, "R2"],
      F = fit[factor_name, "F"],
      p = fit[factor_name, "Pr(>F)"]
    )
  }

  # Combining and BH-adjustment within current factor
  out <- do.call(rbind, results)
  out$p_adj <- p.adjust(out$p, method = adjust)
  rownames(out) <- NULL
  out
}

# Helper function for pairwise PERMDISP comparisons ----------------------------

# Adapted from:
# https://gist.github.com/mcgoodman/58c9d1257fd1625954a4ffa1c3301939
# And: the pairwise.adonis2() function
# Using ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
pairwise_permdisp <- function(dist, factor_name, meta,
                              permutations = 999, adjust = "BH",
                              seed = 122863) {

  # Aligning metadata rows to the distance object
  meta <- meta[labels(dist), , drop = FALSE]
  dmat <- as.matrix(dist)

  # Factor vector and all level combination
  fac <- droplevels(as.factor(meta[[factor_name]]))
  pairs <- combn(levels(fac), 2, simplify = FALSE)

  # Initializing a results list
  results <- list()

  # Iterating through the possible combinations, calculating the statistics
  # for each variable restricted to a pair of values
  for (pr in pairs) {

    # Subsetting to the two levels being compared
    keep  <- fac %in% pr
    d_sub <- as.dist(dmat[keep, keep])
    g_sub <- droplevels(fac[keep])

    # Dispersion test - no covariates
    set.seed(seed)
    bd <- betadisper(d_sub, g_sub)
    pt <- permutest(bd, permutations = permutations)

    # Tidy results
    results[[paste(pr, collapse = "_vs_")]] <- data.frame(
      comparison = paste(pr, collapse = "_vs_"),
      n1 = sum(g_sub == pr[1]),
      n2 = sum(g_sub == pr[2]),
      F  = pt$tab["Groups", "F"],
      p  = pt$tab["Groups", "Pr(>F)"]
    )
  }

  # Combining and BH-adjustment within current factor
  out <- do.call(rbind, results)
  out$p_adj <- p.adjust(out$p, method = adjust)
  rownames(out) <- NULL
  out
}
```

### Bray-Curtis Dissimilarity

```{r bray_curtis}
# PERMANOVA
if (file.exists("avg_results_set2/permanova_bray.RDS")) {
  permanova_bray <- readRDS("avg_results_set2/permanova_bray.RDS")
} else {
  set.seed(122863)
  permanova_bray <- adonis2(avg_bray ~ sex + breed_clean + neutered +
                              active + body_condition + diet + biotics +
                              gastro + skin + neuro + healthy + worming +
                              antibiotics + location + age_categorical + batch,
                            data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_bray, file = "avg_results_set2/permanova_bray.RDS")
}

# PERMDISP
if (file.exists("avg_results_set2/disp_bray.RDS")) {
  disp_bray <- readRDS("avg_results_set2/disp_bray.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_bray), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_bray), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_bray <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_bray,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_bray, file = "avg_results_set2/disp_bray.RDS")
}
```

### Jaccard Distance

```{r jaccard}
# PERMANOVA
if (file.exists("avg_results_set2/permanova_jac.RDS")) {
  permanova_jac <- readRDS("avg_results_set2/permanova_jac.RDS")
} else {
  set.seed(122863)
  permanova_jac <- adonis2(avg_jac ~ sex + breed_clean + neutered +
                             active + body_condition + diet + biotics +
                             gastro + skin + neuro + healthy + worming +
                             antibiotics + location + age_categorical + batch,
                           data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_jac, file = "avg_results_set2/permanova_jac.RDS")
}

# PERMDISP
if (file.exists("avg_results_set2/disp_jac.RDS")) {
  disp_jac <- readRDS("avg_results_set2/disp_jac.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_jac), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_jac), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_jac <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_jac,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_jac, file = "avg_results_set2/disp_jac.RDS")
}
```

### Unweighted UniFrace

```{r uwu}
# PERMANOVA
if (file.exists("avg_results_set2/permanova_uwu.RDS")) {
  permanova_uwu <- readRDS("avg_results_set2/permanova_uwu.RDS")
} else {
  set.seed(122863)
  permanova_uwu <- adonis2(avg_uwu ~ sex + breed_clean + neutered +
                             active + body_condition + diet + biotics +
                             gastro + skin + neuro + healthy + worming +
                             antibiotics + location + age_categorical + batch,
                           data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_uwu, file = "avg_results_set2/permanova_uwu.RDS")
}

# PERMDISP
if (file.exists("avg_results_set2/disp_uwu.RDS")) {
  disp_uwu <- readRDS("avg_results_set2/disp_uwu.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_uwu), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_uwu), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_uwu <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_uwu,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_uwu, file = "avg_results_set2/disp_uwu.RDS")
}
```

### Weighted UniFrac

```{r wu}
# PERMANOVA
if (file.exists("avg_results_set2/permanova_wu.RDS")) {
  permanova_wu <- readRDS("avg_results_set2/permanova_wu.RDS")
} else {
  set.seed(122863)
  permanova_wu <- adonis2(avg_wu ~ sex + breed_clean + neutered +
                            active + body_condition + diet + biotics +
                            gastro + skin + neuro + healthy + worming +
                            antibiotics + location + age_categorical + batch,
                          data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_wu, file = "avg_results_set2/permanova_wu.RDS")
}

# PERMDISP
if (file.exists("avg_results_set2/disp_wu.RDS")) {
  disp_wu <- readRDS("avg_results_set2/disp_wu.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_wu), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_wu), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_wu <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_wu,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_wu, file = "avg_results_set2/disp_wu.RDS")
}
```

### Post-hoc Comparisons

Post-hoc pairwise comparisons for the significant multi-level factors.  
Subsetting the data for each pair combination, and performing a PERMANOVA
while controlling for the strongest factors. Additional pairwise PERMDISP is
performed.

```{r beta_posthoc}
# PERMANOVA pairwise comparisons -----------------------------------------------
if (file.exists("avg_results_set2/pw_permanova.RDS")) {
  pw_permanova <- readRDS("avg_results_set2/pw_permanova.RDS")
} else {

  # Helper function to find significant multi-level factors in a given omnibus
  # PERMANOVA
  sig_multilevel <- function(permanova, meta, alpha = 0.05) {
    terms <- setdiff(rownames(permanova), c("Residual", "Total"))

    # Keeping factors present in metadata
    terms <- terms[terms %in% names(meta) &
                     vapply(meta[terms], is.factor, logical(1))]

    # Significant in the omnibus
    pvals <- permanova[terms, "Pr(>F)"]
    terms <- terms[!is.na(pvals) & pvals < alpha]

    # Only w/ more than two levels
    terms[vapply(terms,
                 function(v) nlevels(droplevels(meta[[v]])) > 2,
                 logical(1))]
  }

  # Confounders
  adj <- c("batch", "age_categorical", "antibiotics", "diet")

  # Helper function to run all significant multi-level factors for one metric
  run_metric <- function(dist, permanova) {
    meta_m <- md_avg[labels(dist), , drop = FALSE]
    facs <- sig_multilevel(permanova, meta_m)
    if (!length(facs)) return(NULL)

    res <- lapply(facs, function(f) {
      out <- pairwise_permanova(
        dist = dist,
        factor_name = f,
        # Not adjusting a factor for itself
        covars = setdiff(adj, f),
        meta = md_avg
      )
      out$term <- f
      out
    })
    do.call(rbind, res)
  }

  set.seed(122863)
  pw_permanova <- list(
    bray = run_metric(avg_bray, permanova_bray),
    jac = run_metric(avg_jac, permanova_jac),
    uwu = run_metric(avg_uwu, permanova_uwu),
    wu = run_metric(avg_wu, permanova_wu)
  )

  saveRDS(pw_permanova, "avg_results_set2/pw_permanova.RDS")
}

# PERMDISP pairwise comparisons ------------------------------------------------
if (file.exists("avg_results_set2/pw_permdisp.RDS")) {
  pw_permdisp <- readRDS("avg_results_set2/pw_permdisp.RDS")
} else {

  # Helper function to find significant multi-level factors in a given omnibus
  # PERMANOVA
  sig_multilevel <- function(permanova, meta, alpha = 0.05) {
    terms <- setdiff(rownames(permanova), c("Residual", "Total"))

    # Keeping factors present in metadata
    terms <- terms[terms %in% names(meta) &
                     vapply(meta[terms], is.factor, logical(1))]

    # Significant in the omnibus
    pvals <- permanova[terms, "Pr(>F)"]
    terms <- terms[!is.na(pvals) & pvals < alpha]

    # Only w/ more than two levels
    terms[vapply(terms,
                 function(v) nlevels(droplevels(meta[[v]])) > 2,
                 logical(1))]
  }

  # Helper function to run all significant multi-level factors for one metric
  # (dispersion takes no covariates)
  disp_metric <- function(dist, permanova) {
    meta_m <- md_avg[labels(dist), , drop = FALSE]
    facs <- sig_multilevel(permanova, meta_m)
    if (!length(facs)) return(NULL)

    res <- lapply(facs, function(f) {
      out <- pairwise_permdisp(
        dist = dist,
        factor_name = f,
        meta = md_avg
      )
      out$term <- f
      out
    })
    do.call(rbind, res)
  }

  set.seed(122863)
  pw_permdisp <- list(
    bray = disp_metric(avg_bray, permanova_bray),
    jac = disp_metric(avg_jac, permanova_jac),
    uwu = disp_metric(avg_uwu, permanova_uwu),
    wu = disp_metric(avg_wu, permanova_wu)
  )

  saveRDS(pw_permdisp, "avg_results_set2/pw_permdisp.RDS")
}
```

### Plots per Variable

Plots and summary statistics for significant covariates.

```{r ordination_plots}
# Ordination for plots
set.seed(122863)
bray_ord <- ordinate(ps_f, method = "PCoA", distance = avg_bray)
set.seed(122863)
jac_ord <- ordinate(ps_f, method = "PCoA", distance = avg_jac)
set.seed(122863)
uwu_ord <- ordinate(ps_f, method = "PCoA", distance = avg_uwu)
set.seed(122863)
wu_ord <- ordinate(ps_f, method = "PCoA", distance = avg_wu)
```

#### Sex

```{r beta_sex}
beta_sex <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "sex", shape = "sex") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = sex, lty = sex)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Sex", shape = "Sex",
         linetype = "Sex", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "sex", shape = "sex") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = sex, lty = sex)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Sex", shape = "Sex",
         linetype = "Sex", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "sex", shape = "sex") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = sex, lty = sex)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Sex", shape = "Sex",
         linetype = "Sex", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "sex", shape = "sex") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = sex, lty = sex)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Sex", shape = "Sex",
         linetype = "Sex", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_sex <- annotate_figure(beta_sex,
                            top = text_grob(paste0("Sex - Beta ",
                                                   "Diversity Metrics"),
                                            face = "bold", size = 14))
```

#### Breed

```{r beta_breed}
# TODO: need to think of plotting
```

#### Physical Activity

```{r beta_active}
beta_active <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "active", shape = "active") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = active, lty = active)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Physical Activity", shape = "Physical Activity",
         linetype = "Physical Activity", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "active", shape = "active") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = active, lty = active)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Physical Activity", shape = "Physical Activity",
         linetype = "Physical Activity", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "active", shape = "active") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = active, lty = active)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Physical Activity", shape = "Physical Activity",
         linetype = "Physical Activity", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "active", shape = "active") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = active, lty = active)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Physical Activity", shape = "Physical Activity",
         linetype = "Physical Activity", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_active <- annotate_figure(beta_active,
                               top = text_grob(paste0("Physical Activity - ",
                                                      "Beta ",
                                                      "Diversity Metrics"),
                                               face = "bold", size = 14))
```

#### Body Condition

```{r beta_body_condition}
beta_body_condition <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord,
                  color = "body_condition", shape = "body_condition") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = body_condition, lty = body_condition)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Body Condition", shape = "Body Condition",
         linetype = "Body Condition", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord,
                  color = "body_condition", shape = "body_condition") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = body_condition, lty = body_condition)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Body Condition", shape = "Body Condition",
         linetype = "Body Condition", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord,
                  color = "body_condition", shape = "body_condition") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = body_condition, lty = body_condition)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Body Condition", shape = "Body Condition",
         linetype = "Body Condition", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord,
                  color = "body_condition", shape = "body_condition") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = body_condition, lty = body_condition)) +
    scale_linetype_manual(values = c(1, 2, 4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Body Condition", shape = "Body Condition",
         linetype = "Body Condition", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_body_condition <- annotate_figure(beta_body_condition,
                                       top = text_grob(paste0("Body Condition ",
                                                              "- Beta ",
                                                              "Diversity ",
                                                              "Metrics"),
                                                       face = "bold",
                                                       size = 14))
```

#### Diet

```{r beta_diet}
beta_diet <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "diet", shape = "diet") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = diet, lty = diet)) +
    scale_linetype_manual(values = c(1:6)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Diet", shape = "Diet",
         linetype = "Diet", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "diet", shape = "diet") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = diet, lty = diet)) +
    scale_linetype_manual(values = c(1:6)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Diet", shape = "Diet",
         linetype = "Diet", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "diet", shape = "diet") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = diet, lty = diet)) +
    scale_linetype_manual(values = c(1:6)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Diet", shape = "Diet",
         linetype = "Diet", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "diet", shape = "diet") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = diet, lty = diet)) +
    scale_linetype_manual(values = c(1:6)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Diet", shape = "Diet",
         linetype = "Diet", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_diet <- annotate_figure(beta_diet,
                             top = text_grob(paste0("Diet - ",
                                                    "Beta ",
                                                    "Diversity Metrics"),
                                             face = "bold", size = 14))
```

#### Biotics

```{r beta_biotics}
beta_biotics <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "biotics", shape = "biotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = biotics, lty = biotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Biotics", shape = "Biotics",
         linetype = "Biotics", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "biotics", shape = "biotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = biotics, lty = biotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Biotics", shape = "Biotics",
         linetype = "Biotics", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "biotics", shape = "biotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = biotics, lty = biotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Biotics", shape = "Biotics",
         linetype = "Biotics", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "biotics", shape = "biotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = biotics, lty = biotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Biotics", shape = "Biotics",
         linetype = "Biotics", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_biotics <- annotate_figure(beta_biotics,
                                top = text_grob(paste0("Biotics - ",
                                                       "Beta ",
                                                       "Diversity Metrics"),
                                                face = "bold", size = 14))
```

#### Skin-related Symptoms

```{r beta_skin}
beta_skin <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "skin", shape = "skin") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = skin, lty = skin)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Skin-related Symptoms", shape = "Skin-related Symptoms",
         linetype = "Skin-related Symptoms", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "skin", shape = "skin") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = skin, lty = skin)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Skin-related Symptoms", shape = "Skin-related Symptoms",
         linetype = "Skin-related Symptoms", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "skin", shape = "skin") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = skin, lty = skin)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Skin-related Symptoms", shape = "Skin-related Symptoms",
         linetype = "Skin-related Symptoms",
         caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "skin", shape = "skin") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = skin, lty = skin)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Skin-related Symptoms", shape = "Skin-related Symptoms",
         linetype = "Skin-related Symptoms", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_skin <- annotate_figure(beta_skin,
                             top = text_grob(paste0("Skin-related ",
                                                    "Symptoms - Beta ",
                                                    "Diversity Metrics"),
                                             face = "bold", size = 14))
```

#### Health

```{r beta_healthy}
beta_healthy <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "healthy", shape = "healthy") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = healthy, lty = healthy)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Healthy", shape = "Healthy",
         linetype = "Healthy", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "healthy", shape = "healthy") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = healthy, lty = healthy)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Healthy", shape = "Healthy",
         linetype = "Healthy", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "healthy", shape = "healthy") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = healthy, lty = healthy)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Healthy", shape = "Healthy",
         linetype = "Healthy", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "healthy", shape = "healthy") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = healthy, lty = healthy)) +
    scale_linetype_manual(values = c(1:2)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # Longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Healthy", shape = "Healthy",
         linetype = "Healthy", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_healthy <- annotate_figure(beta_healthy,
                                top = text_grob(paste0("Health - ",
                                                       "Beta ",
                                                       "Diversity Metrics"),
                                                face = "bold", size = 14))
```

#### Antibiotics

```{r beta_antibiotics}
beta_antibiotics <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord,
                  color = "antibiotics", shape = "antibiotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = antibiotics, lty = antibiotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Antibiotic Usage", shape = "Antibiotic Usage",
         linetype = "Antibiotic Usage", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "antibiotics", shape = "antibiotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = antibiotics, lty = antibiotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Antibiotic Usage", shape = "Antibiotic Usage",
         linetype = "Antibiotic Usage", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "antibiotics", shape = "antibiotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = antibiotics, lty = antibiotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Antibiotic Usage", shape = "Antibiotic Usage",
         linetype = "Antibiotic Usage", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "antibiotics", shape = "antibiotics") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = antibiotics, lty = antibiotics), linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Antibiotic Usage", shape = "Antibiotic Usage",
         linetype = "Antibiotic Usage", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_antibiotics <- annotate_figure(beta_antibiotics,
                                    top = text_grob(paste0("Antibiotic Usage ",
                                                           "- Beta ",
                                                           "Diversity Metrics"),
                                                    face = "bold", size = 14))
```

#### Age

```{r beta_age}
beta_age <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord,
                  color = "age_categorical", shape = "age_categorical") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = age_categorical,
                     lty = age_categorical),
                 linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Age", shape = "Age",
         linetype = "Age", caption = "Bray-Curtis PCoA"),

  # Jaccard
  plot_ordination(ps_f, jac_ord,
                  color = "age_categorical", shape = "age_categorical") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = age_categorical,
                     lty = age_categorical),
                 linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Age", shape = "Age",
         linetype = "Age", caption = "Jaccard (binary) PCoA"),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord,
                  color = "age_categorical", shape = "age_categorical") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = age_categorical,
                     lty = age_categorical),
                 linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Age", shape = "Age",
         linetype = "Age", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord,
                  color = "age_categorical", shape = "age_categorical") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = age_categorical,
                     lty = age_categorical),
                 linewidth = 0.8) +
    scale_linetype_manual(values = c(1:4)) +
    # Bigger shape in legend
    guides(shape = guide_legend(override.aes = list(size = 4))) +
    # longer line in legend
    theme(legend.key.width = unit(2.5, "cm")) +
    labs(color = "Age", shape = "Age",
         linetype = "Age", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_age <- annotate_figure(beta_age,
                            top = text_grob(paste0("Age - ",
                                                   "Beta ",
                                                   "Diversity Metrics"),
                                            face = "bold", size = 14))
```

#### Batch

```{r beta_batch}
beta_batch <- ggarrange(
  # Bray-Curtis
  plot_ordination(ps_f, bray_ord, color = "batch") +
    geom_point(size = 1) +
    labs(color = "Batch", caption = "Bray-Curtis PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  # Jaccard
  plot_ordination(ps_f, jac_ord, color = "batch") +
    geom_point(size = 1) +
    labs(color = "Batch", caption = "Jaccard (binary) PCoA") +
    stat_ellipse(aes(group = batch), linetype = 2),

  # Unweighted UniFrac
  plot_ordination(ps_f, uwu_ord, color = "batch") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Batch", caption = "Unweighted UniFrac PCoA"),

  # Weighted UniFrac
  plot_ordination(ps_f, wu_ord, color = "batch") +
    geom_point(size = 1) +
    stat_ellipse(aes(group = batch), linetype = 2) +
    labs(color = "Batch", caption = "Weighted UniFrac PCoA"),

  ncol = 2, nrow = 2,
  common.legend = TRUE, legend = "bottom", align = "hv"
)
beta_batch <- annotate_figure(beta_batch,
                              top = text_grob(paste0("Batch - ",
                                                     "Beta ",
                                                     "Diversity Metrics"),
                                              face = "bold", size = 14))
```

## Taxonomy and Abundance

Extracting relative abundance and calculating occurrence by taxonomic ranks.

```{r abundance}
# Relative abundance
ps_rel <- microbiome::transform(ps_f, "compositional")
taxon_table <- psmelt(ps_rel)
# TODO: perhaps look at the healthy subset
```

### Phylum

```{r phylum}
# Retrieving phylum-level information
# Mean taxa across samples
tax_table_phy <- taxon_table %>%

  # Handling missing data
  mutate(
    Phylum = as.character(Phylum),
    Phylum = trimws(Phylum),
    Phylum = na_if(Phylum, ""),
    Phylum = replace_na(Phylum, "Unclassified")
  ) %>%

  # Phylum-abundance per sample
  group_by(Sample, Phylum) %>%
  summarize(sample_phylum_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of phyla across samples
  group_by(Phylum) %>%
  summarize(

    # Number of samples with phyla
    num_samples = sum(sample_phylum_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_phylum_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_phylum_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_phy$num_samples == 1)

# Top phyla and unclassified
head(tax_table_phy)
tax_table_phy[tax_table_phy$Phylum == "Unclassified", ]
```

### Class

```{r class}
# Retrieving class-level information
# Mean taxa across samples
tax_table_class <- taxon_table %>%

  # Handling missing data
  mutate(
    Class = as.character(Class),
    Class = trimws(Class),
    Class = na_if(Class, ""),
    Class = replace_na(Class, "Unclassified")
  ) %>%

  # Class-abundance per sample
  group_by(Sample, Class) %>%
  summarize(sample_class_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of class across samples
  group_by(Class) %>%
  summarize(

    # Number of samples with class
    num_samples = sum(sample_class_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_class_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_class_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_class$num_samples == 1)

# Top classes and unclassified
head(tax_table_class)
tax_table_class[tax_table_class$Class == "Unclassified", ]
```

### Order

```{r order}
# Retrieving order-level information
# Mean taxa across samples
tax_table_order <- taxon_table %>%

  # Handling missing data
  mutate(
    Order = as.character(Order),
    Order = trimws(Order),
    Order = na_if(Order, ""),
    Order = replace_na(Order, "Unclassified")
  ) %>%

  # Order-abundance per sample
  group_by(Sample, Order) %>%
  summarize(sample_order_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of order across samples
  group_by(Order) %>%
  summarize(

    # Number of samples with order
    num_samples = sum(sample_order_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_order_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_order_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_order$num_samples == 1)

# Top orders and unclassified
head(tax_table_order)
tax_table_order[tax_table_order$Order == "Unclassified", ]
```

### Family

```{r family}
# Retrieving family-level information
# Mean taxa across samples
tax_table_family <- taxon_table %>%

  # Handling missing data
  mutate(
    Family = as.character(Family),
    Family = trimws(Family),
    Family = na_if(Family, ""),
    Family = replace_na(Family, "Unclassified")
  ) %>%

  # Family-abundance per sample
  group_by(Sample, Family) %>%
  summarize(sample_family_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of family across samples
  group_by(Family) %>%
  summarize(

    # Number of samples with family
    num_samples = sum(sample_family_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_family_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_family_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_family$num_samples == 1)

# Top families and unclassified
head(tax_table_family)
tax_table_family[tax_table_family$Family == "Unclassified", ]
```

### Genus

```{r genus}
# Retrieving genus-level information
# Mean taxa across samples
tax_table_genus <- taxon_table %>%

  # Handling missing data
  mutate(
    Genus = as.character(Genus),
    Genus = trimws(Genus),
    Genus = na_if(Genus, ""),
    Genus = replace_na(Genus, "Unclassified")
  ) %>%

  # Genus-abundance per sample
  group_by(Sample, Genus) %>%
  summarize(sample_genus_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of genus across samples
  group_by(Genus) %>%
  summarize(

    # Number of samples with genus
    num_samples = sum(sample_genus_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_genus_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_genus_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_genus$num_samples == 1)

# Top genera and unclassified
head(tax_table_genus)
tax_table_genus[tax_table_genus$Genus == "Unclassified", ]
```

### Differential Abundance

Performed in full on the current pipeline only. 

```{r prevalent_abundances}
# Checking prevalence threshold
# Helper function to calculate prevalence
get_prevalence <- function(taxon) {
  prevalence <- sum(taxon > 0) / length(taxon)
  prevalence
}

# Non-rarefied data
ps_prev <- filter_taxa(ps_f,
                       function(taxon) get_prevalence(taxon) > 0.1,
                       prune = TRUE)

# Some checks
nsamples(ps_prev)
ntaxa(ps_prev) # 210 taxa remaining
summary(sample_sums(ps_prev))
```

#### Models

```{r differential_abundance}
# According to the tutorial:
# https://github.com/biobakery/biobakery/wiki/MaAsLin3
# And manual:
# www.bioconductor.org/packages//release/bioc/vignettes/maaslin3/inst/doc/maaslin3_manual.html

if (file.exists("maaslin_output_set2/da_model.RDS")) {
  da_model <- readRDS("maaslin_output_set2/da_model.RDS")
  da_group_model <- readRDS("maaslin_output_group_set2/da_group_model.RDS")
} else {

  # Count data pre-rarefaction
  otu_da <- as(otu_table(ps_f), "matrix")
  if (taxa_are_rows(ps_f)) otu_da <- t(otu_da)
  otu_da <- data.frame(otu_da)

  # Appropriate metadata
  md_da <- data.frame(sample_data(ps_f))

  # Adding recommended sampling depth
  md_da$read_depth <- sample_sums(ps_f)[rownames(md_da)]

  # Running MaAsLin 3
  # Selecting known fixed effects and variables with significant associations in
  # previous analysis. Logic according to:
  # https://forum.biobakery.org/t/selecting-covariates-maaslin2-da/8887
  set.seed(122863)
  da_model <- maaslin3(
    input_data = otu_da,
    input_metadata = md_da,
    output = "maaslin_output_set2",
    formula = "~ sex + group(breed_clean) + active + body_condition + diet +
                      biotics + gastro + skin + healthy + antibiotics +
                      age_categorical + read_depth + group(batch)",
    normalization = "TSS",
    transform = "LOG",
    augment = TRUE,
    max_significance = 0.1,
    min_prevalence = 0.1
  )
  saveRDS(da_model, file = "maaslin_output_set2/da_model.RDS")
}
```

#### Convergence

```{r da_convergence}
main <- read.delim("maaslin_output_set2/all_results.tsv", check.names = FALSE)
main$feature <- sub("^X", "", main$feature)

# Number and type of errors
sum(!is.na(main$error)) # 148
sum(!is.na(main$error)) * 100 / nrow(main) # error 1.41%, success 98.59%
table(main$error) # Fitting success - 99.91%

# Checking if failures are concentrated on one variable
table(main$metadata[!is.na(main$error)])

# Abundance vs. prevalence failures
table(main$model[!is.na(main$error)])
```

#### Statistically Signficant Features

```{r sig_da}
sig <- read.delim("maaslin_output_set2/significant_results.tsv",
                  check.names = FALSE)
sig$feature <- sub("^X", "", sig$feature)

# Total significant associations (qval_joint) - 1249
nrow(sig)

# Individually significant models - 689
sum(sig$qval_individual < 0.1, na.rm = TRUE)

# Unique feature and metadata (covariate level for non-group) associations - 637
sig %>%
  distinct(feature, metadata, value) %>%
  nrow()

# Unique significant features - 192
length(unique(sig$feature))

# Abundance vs. prevalence
table(sig$model)

# Significance per covariate per model
table(sig$metadata, sig$model)

# Per covariate only
sig %>%
  distinct(feature, metadata) %>%
  count(metadata, sort = TRUE)
```

#### Summary and Plotting

Summary tables per model type and covariate, divided by levels. Including
coefficient, and p- and q-values.

```{r da_summary}
# Debugged with ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
## Readable labels
# Helper function for table and plot labels
make_labels <- function(features, tax_mat) {
  ranks <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  tx <- as.data.frame(tax_mat[features, , drop = FALSE],
                      stringsAsFactors = FALSE)
  tx$feature <- rownames(tx)

  # Cleaning function
  clean <- function(x) {
    x <- str_replace(x, "^[a-z]__", "")
    x <- trimws(x)
    x[x == ""] <- NA
    x
  }
  tx[ranks] <- lapply(tx[ranks], clean)

  # Finest taxonomic rank
  finest <- apply(tx[ranks], 1,
                  function(r) {
                    r <- r[!is.na(r)]
                    if (!length(r)) "Unclassified" else tail(r, 1)
                  })

  # Extracting phylum
  phylum <- ifelse(is.na(tx$Phylum), "Unclassified", tx$Phylum)

  # Short and longer labels
  tibble(feature = tx$feature, phylum = phylum, finest = finest) %>%

    # Consistent labeling
    arrange(phylum, finest, feature) %>%
    group_by(phylum) %>%
    mutate(idx = row_number(),
           label = paste0(phylum, " ", idx),
           label_rich  = paste0(finest, " (", phylum, " ", idx, ")")) %>%
    ungroup()
}

# Summary tables
main <- main[is.na(main$error), ]

# Unique labeling for all features
all_feats <- unique(main$feature)
global_labels <- make_labels(all_feats, tax_mat)

# Helper function to generate tables per covariate
# Takes the results table, covariate model type (prevalence/abundance), labels,
# significance choice and a delimiter for sub headers
cov_table <- function(df, covariate, model_type, labels,
                      sig_only = TRUE, delim = "_") {

  # Filtering per covariate and model type
  sub <- df %>% filter(metadata == covariate, model == model_type)

  # Exiting for no results
  if (!nrow(sub)) return(NULL)

  # Subsetting relevant information
  sub <- sub %>%
    left_join(labels %>%
                select(feature, label_rich),
              by = "feature") %>%
    rename(level = value, coef = coef,
           p = pval_individual, q = qval_individual)

  # Only significant results
  if (sig_only) {
    keep <- sub %>%
      group_by(feature) %>%
      summarize(minq = min(q, na.rm = TRUE), .groups = "drop") %>%
      filter(minq < 0.1) %>%
      pull(feature)
    sub <- filter(sub, feature %in% keep)
    if (!nrow(sub)) return(NULL)
  }

  # Wide format for readability
  wide <- sub %>%
    select(taxon = label_rich, level, coef, p, q) %>%
    pivot_wider(id_cols = taxon,
                names_from = level,
                values_from = c(coef, p, q),
                names_glue = paste0("{level}", delim, "{.value}"))

  # Ordering so the three sub-columns are together per level
  lvls <- unique(sub$level)
  ord  <- as.vector(t(outer(lvls, c("coef", "p", "q"),
                            function(a, b) paste0(a, delim, b))))
  wide %>% select(taxon, any_of(ord)) %>% arrange(taxon)
}

# Wanted covariates
covariates <- c("sex", "breed_clean", "active", "diet", "biotics",
                "gastro", "skin", "healthy", "antibiotics", "age_categorical")

# List per model
tables_abund <- setNames(lapply(covariates, cov_table,
                                df = main, model_type = "abundance",
                                labels = global_labels), covariates)
tables_prev  <- setNames(lapply(covariates, cov_table,
                                df = main, model_type = "prevalence",
                                labels = global_labels), covariates)

# Dropping covariates with no significant features
tables_abund <- tables_abund[!sapply(tables_abund, is.null)]
tables_prev  <- tables_prev [!sapply(tables_prev, is.null)]
```

Heatmap plotting the top 25 features (by q-value) across models, including
significance estimate. Additionally trying to build a single table of the top
features associated by covariate. 

```{r da_plot}
# Debugged with ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
# Removing "uninteresting" covariates
bio <- main %>%
  filter(!metadata %in% c("breed_clean", "batch", "read_depth"), is.na(error))

# Picking the top features by strongest association (min q across rows)
top_feats <- bio %>%
  group_by(feature) %>%
  summarize(minq = min(qval_individual, na.rm = TRUE, .groups = "drop")) %>%
  arrange(minq) %>%
  slice_head(n = 25) %>%
  pull(feature)

# Data for plotting
hm <- bio %>%
  filter(feature %in% top_feats) %>%
  left_join(global_labels %>%
              select(feature, label, phylum),
            by = "feature") %>%
  mutate(term = ifelse(is.na(value) | value == "",
                       metadata,
                       paste0(metadata, ": ", value)),

         # Significance symbol by q-value
         star = cut(qval_individual,
                    c(-Inf, 0.001, 0.01, 0.1, Inf),
                    labels = c("***", "**", "*", "")))

# Heatmap with ggplot
# https://r-graph-gallery.com/79-levelplot-with-ggplot2.html
da_heatmap <- ggplot(hm,
                     aes(x = term,
                         y = reorder(label, coef),
                         fill = coef)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = star), size = 3) +
  # Grouping taxa by phylum and model
  facet_grid(phylum ~ model, scales = "free", space = "free") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, name = "Coef") +
  labs(x = NULL, y = NULL,
       title = "Differential Abundance - Top 25 Features",
       caption = paste0("Abundance coef = log-scale; Prevalence coef = ",
                        "log-odds; * q < 0.1, ** q < 0.01, *** q < 0.001")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        strip.text.y = element_text(angle = 0),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.caption = element_text(size = 12))

# Significant data for summary
sig_bio <- bio %>%
  filter(qval_individual < 0.1)

# Summary table
da_top_table <- sig_bio %>%

  # Filtering for top features
  filter(feature %in% top_feats) %>%

  # Nice labels
  left_join(global_labels %>%
              select(feature, label_rich), by = "feature") %>%
  transmute(
    taxon = label_rich,
    covariate = metadata,

    # Level vs. reference
    comparison = value,
    model = model,
    coef = round(coef, 4),
    q = signif(qval_individual, 4)
  ) %>%
  arrange(covariate, model, q)
```
