---
title: "Dissertation Statistical Analysis - OTU (de novo)"
author: "B286413"
date: "2026-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Downloading unavailable packages
# BioManager
#install.packages("BiocManager", repos = "https://cloud.r-project.org")
# phyloseq
#BiocManager::install("phyloseq", ask = FALSE, update = FALSE)
# QIIME2
#install.packages("remotes", repos = "https://cloud.r-project.org")
#remotes::install_github("jbisanz/qiime2R")
# picante
#BiocManager::install("picante", ask = FALSE, update = FALSE)
# ggpubr
#install.packages("ggpubr")
# microbiome
#BiocManager::install("microbiome")
# lmerTest
#install.packages("lmerTest")
#DHARMa
#install.packages("DHARMa")
# Masslin2
#BiocManager::install("Maaslin2")
# MaAsLin 3
#BiocManager::install("biobakery/maaslin3")
# mia
#BiocManager::install("mia")
# pairwiseAdonis
#remotes::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
# FSA - Dunn's test
#install.packages("FSA")
# effsize - Cliff's delta, Dunn's effect size
#install.packages("effsize")

# Loading packages
library(tidyverse)
library(phyloseq)
library(qiime2R)
library(vegan)
library(picante)
library(ggpubr)
library(microbiome)
library(lme4)
library(MASS)
library(DHARMa)
library(Maaslin2)
library(maaslin3)
library(mia)
library(emmeans)
library(pairwiseAdonis)
library(FSA)
library(effsize)

# Resolving conflicts
library(conflicted)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::count)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::desc)
conflicts_prefer(base::intersect)
conflicts_prefer(dplyr::inner_join)
conflicts_prefer(base::summary)
conflicts_prefer(dplyr::left_join)
conflicts_prefer(dplyr::rename)
conflicts_prefer(base::as.factor)
conflicts_prefer(base::unname)

# Working directory
setwd("~/project")
```

## Loading Data

### Data Description

* The data was generated following a standard QIIME2 analysis workflow.
* ./otu_otus_set2/otu97_table_lenfilt.qza - merged feature table
* .metadata/metadata_clean.rds - clean metadata file
* ./otu_otus_set2/otu97_taxonomy.qza - taxonomic classification file
* ./phylogeny_otu_set2/otu97_rooted.qza - rooted phylogenetic tree

```{r loading_data}
## Paths
path_table <- "./otu_otus_set2/otu97_table_lenfilt.qza"
path_tax <- "./otu_otus_set2/otu97_taxonomy.qza"
path_tree <- "./phylogeny_otu_set2/otu97_rooted.qza"
path_md_rds <- "./metadata/metadata_clean.rds"

# Feature table
features <- read_qza(path_table)$data
colnames(features) <- trimws(as.character(colnames(features)))

# Metadata
md <- readRDS(path_md_rds)
md <- as.data.frame(md)

# Cleaning sample IDs
colnames(features) <- trimws(as.character(colnames(features)))
md$activation_code <- trimws(as.character(md$activation_code))
md <- md[!is.na(md$activation_code) & md$activation_code != "", , drop = FALSE]
rownames(md) <- md$activation_code

# Checking duplicates
stopifnot(anyDuplicated(colnames(features)) == 0)
stopifnot(anyDuplicated(rownames(md)) == 0)

# Subsetting to metadata overlap, keeping feature-table order
keep_samples <- colnames(features)[colnames(features) %in% rownames(md)]
features <- features[, keep_samples, drop = FALSE]

# Reordering metadata to match features columns
md <- md[colnames(features), , drop = FALSE]

# Sanity check
stopifnot(identical(colnames(features), rownames(md)))

# Removing absent taxa after subsetting
keep_taxa_table <- rownames(features)[rowSums(features) > 0]
features <- features[keep_taxa_table, , drop = FALSE]

# Taxonomy
tax_df <- read_qza(path_tax)$data
# Updating row names
rownames(tax_df) <- as.character(tax_df$Feature.ID)
# Converting taxonomy into a rank matrix for phyloseq
tax_mat <- str_split_fixed(as.character(tax_df$Taxon), ";\\s*", n = 7)
colnames(tax_mat) <- c("Domain", "Phylum", "Class", "Order",
                       "Family", "Genus", "Species")
rownames(tax_mat) <- rownames(tax_df)

# Phylogenetic tree
tree_phylo <- read_qza(path_tree)$data

# Intersection across all components
common_taxa <- Reduce(intersect, list(
  rownames(features),
  rownames(tax_mat),
  tree_phylo$tip.label
))

# Pruning all components to shared taxa
features2 <- features[common_taxa, , drop = FALSE]
tax_mat2 <- tax_mat[common_taxa, , drop = FALSE]
tree2_phylo <- ape::keep.tip(tree_phylo, common_taxa)

# phyloseq components
otu <- otu_table(as.matrix(features2), taxa_are_rows = TRUE)
tax <- tax_table(tax_mat2)
tr <- phy_tree(tree2_phylo)
smd <- sample_data(md)

# phyloseq object
ps <- phyloseq(otu, tax, tr, smd)

# Verifying object - sanity checks
ps
validObject(ps)
nsamples(ps)
ntaxa(ps)

# Depth statistics
summary(sample_sums(ps))
summary(taxa_sums(ps))

# Zero depth samples
sum(sample_sums(ps) == 0)

# Initial depth vs. richness inspection
sample_data(ps)$depth <- sample_sums(ps)
p <- plot_richness(ps, x = "depth", measures = "Observed") +
  scale_x_log10() +
  ggtitle("Observed Richness vs. Depth (pre-filtering)") +
  ylab("Observed Richness") +
  xlab("Sample Depth") +
  theme_bw() +
  theme(
    axis.title.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(face = "bold", size = 10),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
p

# Lowest-depth samples
sort(sample_sums(ps))[1:20]

# Highest richness samples
alpha <- estimate_richness(ps, measures = "Observed")
rownames(alpha)[order(alpha$Observed, decreasing = TRUE)][1:20]
alpha[order(alpha$Observed, decreasing = TRUE),
      "Observed", drop = FALSE][1:20, ]
```

## Filtering Data

Initial inspection in QIIME2 revealed that the raw data contains mitochondrial
and chloroplast taxa, which should be removed, along with eukaryotic sequences.
This part will also remove irregular samples that were previously identified.

```{r filtering}
# Pre-filtering object dimensions
nsamples(ps)
ntaxa(ps)
summary(sample_sums(ps))

# Metadata
md_f <- as(sample_data(ps), "data.frame")

# Duplicated samples based on name and dob -------------------------------------
# Seven remaining due to batch exclusion (21_May_2026)
dup_keys <- md_f %>%
  count(name, dob, name = "n") %>%
  filter(n > 1) %>%
  arrange(desc(n))
dup_keys

# Duplicated rows for metadata inspection
dup_rows <- md_f %>%
  inner_join(dup_keys, by = c("name", "dob")) %>%
  arrange(name, dob)
dup_rows

# Duplicated samples to remove based on metadata
dups <- c(
  "TW59F9", # Almost identical, fewer samples in the other batch
  "KHG2HC", # Fewer samples in other batch
  "VVYVBT", # Low vs. medium physical activity
  "B9KPQR", # Fewer samples in other batch
  "ZCJCAR", # No preference, picked the first one
  "P93VRB", # Overweight vs. normalweight
  "D2WG8P" # Raw vs. mixed diet
)

# Filtering out "bad" samples --------------------------------------------------
# Missing metadata (NA, e.g., negative age) - four samples
missing_md <- unlist(sample_data(ps)[apply(
  sample_data(ps),
  1,
  function(x) any(is.na(x))
), "activation_code"])

# Irregular samples (writing all, some were already filtered in previous steps)
bad_samples <- c(
  dups,
  missing_md,
  "GER69C", # Irregularly low depth
  "86PGKK", # Irregularly low depth
  "NVTNX3", # Low depth - under 40k
  "E27ADN" # Low depth - under 40k
)

ps_f <- prune_samples(setdiff(sample_names(ps), bad_samples), ps)

# Chloroplast, mitochondria and Eukaryota filtering ----------------------------
# Original occurrences
sum(grepl("mitochondria", as.vector(tax_table(ps_f)[, "Family"]),
          ignore.case = TRUE), na.rm = TRUE) # 148 matches
sum(grepl("chloroplast", as.vector(tax_table(ps_f)[, "Order"]),
          ignore.case = TRUE), na.rm = TRUE) # 82 matches
sum(grepl("eukaryota", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 229 matches
sum(grepl("Unassigned", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) # 387

# Checking if these are concentrated in particular samples
contam_taxa <- taxa_names(ps_f)[
  grepl("eukaryota",
        as.vector(tax_table(ps_f)[, "Domain"]),
        ignore.case = TRUE) |
    grepl("chloroplast",
          as.vector(tax_table(ps_f)[, "Order"]),
          ignore.case = TRUE) |
    grepl("mitochondria",
          as.vector(tax_table(ps_f)[, "Family"]),
          ignore.case = TRUE)
]

ps_contam <- prune_taxa(contam_taxa, ps_f)

# Per-sample summary: contaminant reads, total reads, and proportion
per_sample <- tibble(
  sample = sample_names(ps_f),
  contam_reads = sample_sums(ps_contam)[sample_names(ps_f)],
  total_reads = sample_sums(ps_f),
) %>%
  mutate(contam_prop = contam_reads / total_reads,
         contam_pct  = 100 * contam_prop) %>%
  arrange(desc(contam_pct))

# Worst samples
head(per_sample, 20)
summary(per_sample$contam_pct)

# Some more Eukaryota investigation --------------------------------------------
# Percentage
sum(grepl("eukaryota", as.vector(tax_table(ps_f)[, "Domain"]),
          ignore.case = TRUE), na.rm = TRUE) * 100 / ntaxa(ps_f) # 1.121%

# Singletons
euk_taxa <- taxa_names(ps_f)[grepl("eukaryota",
                                   as.vector(tax_table(ps_f)[, "Domain"]),
                                   ignore.case = TRUE)]
euk_sums <- taxa_sums(prune_taxa(euk_taxa, ps_f))
# 0-309, mean 4.179, median 1
summary(euk_sums)
sum(euk_sums == 1) / length(euk_sums)

# Per-batch investigation
ps_euk <- prune_taxa(euk_taxa, ps_f)

otu_euk <- as(otu_table(ps_euk), "matrix")
if (taxa_are_rows(ps_euk)) {
  otu_euk <- t(otu_euk)
}
batch <- as.character(sample_data(ps_f)$batch)
euk_batch <- data.frame(sample = sample_names(ps_f), batch,
                        euk_reads = rowSums(otu_euk),
                        total = sample_sums(ps_f),
                        euk_feats = rowSums(otu_euk > 0)) %>%
  group_by(batch) %>%
  summarize(n = n(),
            euk_pct = 100 * sum(euk_reads) / sum(total),
            mean_euk_feats = mean(euk_feats),
            .groups = "drop") %>%
  arrange(desc(euk_pct))
euk_batch

# Filtering based on taxa ------------------------------------------------------
ps_f <- subset_taxa(
  ps_f,
  !grepl("mitochondria", Family, ignore.case = TRUE) &
    !grepl("chloroplast", Order, ignore.case = TRUE) &
    !grepl("eukaryota", Domain, ignore.case = TRUE)
)
ps_f <- prune_taxa(taxa_sums(ps_f) > 0, ps_f)

# Checking filtered object
nsamples(ps_f)
ntaxa(ps_f) # 19917
summary(sample_sums(ps_f)) # Min depth - 40607
```

## Rarefaction

### Rarefaction Curves

Different methods of "normalization" can be used before conducting statistical
analyses, rarefaction remains one of the best ones (Schloss, 2024). Since the
metadata required manipulation, rarefaction curves are produced in R (rather
than QIIME2).

```{r rarefaction_curves}
# Help from: https://www.youtube.com/watch?v=_OEdFjc1D9I
# Debugging with ELM (GPT 5.2), https://elm.edina.ac.uk/elm/elm
# OTU table as samples x taxa
otu_mat <- as(otu_table(ps_f), "matrix")
if (taxa_are_rows(ps_f)) otu_mat <- t(otu_mat)

# Batch vector
batch <- as.character(sample_data(ps_f)$batch)
names(batch) <- sample_names(ps_f)

# Align batch to OTU rows
batch <- batch[rownames(otu_mat)]

# Depth grid
step_size <- 1000
depths <- rowSums(otu_mat)

# Verifying 80k is within range
xmax <- max(c(depths, 80000), na.rm = TRUE)
depth_grid <- seq(step_size, as.integer(xmax), by = step_size)

# Compute expected richness at each depth
rich_mat <- suppressWarnings(
  sapply(depth_grid, function(d) vegan::rarefy(otu_mat, sample = d))
)

# Verifying steps as columns and samples as rows
colnames(rich_mat) <- as.character(depth_grid)
rownames(rich_mat) <- rownames(otu_mat)

# Long data frame for ggplot
rc_long <- as.data.frame(rich_mat, check.names = FALSE)
rc_long$sample_id <- rownames(rc_long)
rc_long <- rc_long %>%
  pivot_longer(-sample_id, names_to = "depth", values_to = "observed") %>%
  mutate(
    depth = as.numeric(depth),
    batch = batch[sample_id],
    sample_depth = depths[sample_id]
  )

# Stopping the plot when exceeding threshold (NA instead)
rc_long <- rc_long %>%
  mutate(
    sample_depth = depths[sample_id],
    observed = ifelse(depth > sample_depth, NA_real_, observed)
  )

# Defining panels
batches <- sort(unique(rc_long$batch))

# Global minimum
global_min <- min(rc_long$sample_depth, na.rm = TRUE)

threshold_df <- bind_rows(
  # Minimum within each batch
  rc_long %>%
    distinct(sample_id, batch, sample_depth) %>%
    group_by(batch) %>%
    summarize(threshold = min(sample_depth, na.rm = TRUE), .groups = "drop") %>%
    mutate(threshold_type = "min_in_batch"),

  # Global min replicated for each batch
  tibble(batch = batches,
         threshold_type = "global_min",
         threshold = global_min),

  # Fixed thresholds replicated for each batch
  tibble(batch = batches, threshold_type = "40000", threshold = 40000),
  tibble(batch = batches, threshold_type = "60000", threshold = 60000),
  tibble(batch = batches, threshold_type = "80000", threshold = 80000)
) %>%
  mutate(
    threshold_type = factor(
      threshold_type,
      levels = c("min_in_batch", "global_min",
                 "40000", "60000", "80000"),
      labels = c("Minimum Within Batch", "Global Minimum",
                 "40,000", "60,000", "80,000")
    )
  )

## Plots
rc_long$batch <- factor(rc_long$batch, levels = batches)
threshold_df$batch <- factor(threshold_df$batch, levels = batches)

# Plotting a figure for each batch
for (b in levels(rc_long$batch)) {

  th_b <- threshold_df %>%
    filter(batch == b) %>%
    select(threshold_type, threshold)

  df_b <- rc_long %>%
    filter(batch == b) %>%
    select(sample_id, depth, observed, sample_depth) %>%
    tidyr::crossing(th_b) %>%
    filter(depth <= threshold, depth <= sample_depth)

  p <- ggplot(df_b, aes(x = depth, y = observed, group = sample_id)) +
    geom_line(alpha = 0.3, linewidth = 0.45, color = "black", na.rm = TRUE) +
    geom_vline(aes(xintercept = threshold), linetype = 2, color = "red") +
    facet_wrap(~ threshold_type, nrow = 1, scales = "free_x") +
    theme_bw() +
    labs(
      title = paste0("Rarefaction Curves - Batch ", b),
      x = "Reads",
      y = "Expected observed ASVs"
    ) +
    theme(
      strip.text = element_text(size = 9),
      plot.title = element_text(face = "bold", hjust = 0.5)
    )

  ggsave(
    filename = file.path(paste0("rarefaction_plots_otu_set2/rarefaction_batch_",
                                b, ".png")),
    plot = p,
    width = 14,
    height = 4.5,
    dpi = 300
  )
}

## Statistical summary - minimums
min_depth_by_batch <- data.frame(sample_data(ps_f),
                                 stringsAsFactors = FALSE) %>%
  rownames_to_column("sample_id") %>%
  mutate(
    depth = sample_sums(ps_f),
    batch = as.character(batch)
  ) %>%
  group_by(batch) %>%
  summarize(
    min_reads = min(depth),
    .groups = "drop"
  )
```

## Final Descriptive Statistics

Final descriptive information for samples that will proceed into the statistical
analysis after filtering. Counts and percentages are calculated per categorical
variable. Continuous age statistics were also calculated.

```{r descriptive_stats}
md_f <- sample_data(ps_f)
md_f <- as(md_f, "data.frame")

# Sex --------------------------------------------------------------------------
sex_stats <- md_f %>%
  group_by(sex) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Age --------------------------------------------------------------------------
# Categorical
age_cat <- md_f %>%
  group_by(age_categorical) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Continuous
age_cont <- md_f %>%
  summarize(
    Minimum = min(age_years),
    Maximum = max(age_years),
    Median = median(age_years),
    Mean = mean(age_years),
    SD = sd(age_years)
  )

# Batch ------------------------------------------------------------------------
batch_stats <- md_f %>%
  group_by(batch) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Antibiotics ------------------------------------------------------------------
antibiotics_stats <- md_f %>%
  group_by(antibiotics) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Neutering  Status ------------------------------------------------------------
neutering_stats <- md_f %>%
  group_by(neutered) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Combined w/ sex
# Twice as many intact males than females
neutering_sex_stats <- md_f %>%
  count(neutered, sex, name = "n") %>%
  mutate(perc = 100 * n / sum(n))

# Physical Activity ------------------------------------------------------------
active_stats <- md_f %>%
  group_by(active) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Body Condition ---------------------------------------------------------------
body_condition_stats <- md_f %>%
  group_by(body_condition) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Diet -------------------------------------------------------------------------
diet_stats <- md_f %>%
  group_by(diet) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Breed ------------------------------------------------------------------------
breed_stats <- md_f %>%
  group_by(breed_clean) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Biotics ----------------------------------------------------------------------
biotics_stats <- md_f %>%
  group_by(biotics) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Healthy / Symptoms -----------------------------------------------------------
# Categories are not mutually exclusive
symptoms_stats <- md_f %>%
  summarize(
    healthy = sum(healthy %in% "TRUE"),
    gastrointestinal = sum(gastro %in% "TRUE"),
    skin = sum(skin %in% "TRUE"),
    neurological_behavioral = sum(neuro %in% "TRUE")
  )

# Vaccination ------------------------------------------------------------------
vaccination_stats <- md_f %>%
  group_by(vaccination) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Worming ----------------------------------------------------------------------
worming_stats <- md_f %>%
  group_by(worming) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)

# Location ---------------------------------------------------------------------
location_stats <- md_f %>%
  group_by(location) %>%
  summarize(n = n()) %>%
  mutate(perc = (n / sum(n)) * 100)
```

## Alpha Diversity

First assessment of host factors associations with the microbiome is performed
with alpha diversity metrics (within-sample diversity).

```{r avg_alpha}
# Calculating alpha diversity metrics with repeated subsampling
# From:
# https://bioconductor.posit.co/packages/3.23/bioc/manuals/mia/man/mia.pdf

if (file.exists("avg_results_otu_set2/avg_alpha.RDS")) {
  tse <- readRDS("avg_results_otu_set2/avg_alpha.RDS")
} else {
  # Converting to mia-compatible object
  tse <- convertFromPhyloseq(ps_f)

  # 100 iterations, requesting five metrics
  # Chao1 was initially considered but not included in the final analysis
  # because of:
  # https://pubmed.ncbi.nlm.nih.gov/38869966/
  set.seed(122863)
  tse <- addAlpha(
    tse,
    assay.type = "counts",
    index = c("observed", "chao1", "shannon", "inverse_simpson", "faith"),
    sample = global_min,
    niter = 100
  )

  # Checking and saving object
  head(colData((tse)))
  saveRDS(object = tse,
          file = "avg_results_otu_set2/avg_alpha.RDS")
}

# Metadata Object
md_avg <- as.data.frame(colData(tse))
```

### Observed ASVs

```{r observed}
# Distribution -----------------------------------------------------------------
observed_dist <- ggplot(md_avg, aes(x = observed)) +
  geom_histogram() +
  ggtitle("Richness Distribution") +
  xlab("Observed OTUs") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
observed_stats <- md_avg %>%
  summarize(
    Minimum = min(observed),
    Maximum = max(observed),
    Median = median(observed),
    Mean = mean(observed),
    SD = sd(observed)
  )

# Linear model -----------------------------------------------------------------
observed_lm <- lm(observed ~ sex + breed_clean + neutered +
                    active + body_condition + diet + biotics +
                    gastro + skin + neuro + healthy + worming +
                    antibiotics + location + age_categorical + batch,
                  data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(observed_lm, which = c(1, 2, 4))

# Log-transformation -----------------------------------------------------------
md_avg$observed_log <- log(md_avg$observed)
observed_log <- lm(observed_log ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(observed_log, which = c(1, 2, 4))

# Negative binomial ------------------------------------------------------------
observed_nb <- glm.nb(observed ~ sex + breed_clean + neutered +
                        active + body_condition + diet + biotics +
                        gastro + skin + neuro + healthy + worming +
                        antibiotics + location + age_categorical + batch,
                      data = md_avg)
# Checking model fit
summary(observed_nb)
testDispersion(observed_nb)

simulated_observed_nb <- simulateResiduals(fittedModel = observed_nb,
                                           plot = FALSE)
plot(simulated_observed_nb)
```

### Chao1

```{r chao1, echo=FALSE}
# Excluded because of:
# https://pubmed.ncbi.nlm.nih.gov/38869966/
# Distribution -----------------------------------------------------------------
chao1_dist <- ggplot(md_avg, aes(x = chao1)) +
  geom_histogram() +
  ggtitle("Chao1 Distribution") +
  xlab("Chao1") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
chao1_stats <- md_avg %>%
  summarize(
    Minimum = min(chao1),
    Maximum = max(chao1),
    Median = median(chao1),
    Mean = mean(chao1),
    SD = sd(chao1)
  )

# Linear model -----------------------------------------------------------------
chao1_lm <- lm(chao1 ~ sex + breed_clean + neutered +
                 active + body_condition + diet + biotics +
                 gastro + skin + neuro + healthy + worming +
                 antibiotics + location + age_categorical + batch,
               data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(chao1_lm, which = c(1, 2, 4))

# Gamma GLM --------------------------------------------------------------------
chao1_gamma <- glm(chao1 ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   family = Gamma(link = "log"), data = md_avg)

# Checking model fit
testDispersion(chao1_gamma)
simulated_chao_gamma <- simulateResiduals(fittedModel = chao1_gamma,
                                          plot = FALSE)
plot(simulated_chao_gamma)
```

### Shannon's Index

```{r shannon}
# Distribution -----------------------------------------------------------------
shannon_distribution <- ggplot(md_avg, aes(x = shannon)) +
  geom_histogram() +
  ggtitle("Shannon Diversity Distribution") +
  xlab("Shannon's Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
shannon_stats <- md_avg %>%
  summarize(
    Minimum = min(shannon),
    Maximum = max(shannon),
    Median = median(shannon),
    Mean = mean(shannon),
    SD = sd(shannon)
  )

# Linear model -----------------------------------------------------------------
shannon_lm <- lm(shannon ~ sex + breed_clean + neutered +
                   active + body_condition + diet + biotics +
                   gastro + skin + neuro + healthy + worming +
                   antibiotics + location + age_categorical + batch,
                 data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(shannon_lm, which = c(1, 2, 4))
```

### Simpson's Index

```{r inverse_simpson}
# Distribution -----------------------------------------------------------------
simpson_distribution <- ggplot(md_avg, aes(x = inverse_simpson)) +
  geom_histogram() +
  ggtitle("Inverse Simpson Diversity Distribution") +
  xlab("Inverse Simpson Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
simpson_stats <- md_avg %>%
  summarize(
    Minimum = min(inverse_simpson),
    Maximum = max(inverse_simpson),
    Median = median(inverse_simpson),
    Mean = mean(inverse_simpson),
    SD = sd(inverse_simpson)
  )

# Linear model -----------------------------------------------------------------
simpson_lm <- lm(inverse_simpson ~ sex + breed_clean + neutered +
                   active + body_condition + diet + biotics +
                   gastro + skin + neuro + healthy + worming +
                   antibiotics + location + age_categorical + batch,
                 data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(simpson_lm, which = c(1, 2, 4))

# Square-root Transformation ---------------------------------------------------
md_avg$simpson_sqrt <- sqrt(md_avg$inverse_simpson)
simpson_sqrt <- lm(simpson_sqrt ~ sex + breed_clean + neutered +
                     active + body_condition + diet + biotics +
                     gastro + skin + neuro + healthy + worming +
                     antibiotics + location + age_categorical + batch,
                   data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(simpson_sqrt, which = c(1, 2, 4))

# Transformed distribution -----------------------------------------------------
simpson_sqrt_dist <- ggplot(md_avg, aes(x = simpson_sqrt)) +
  geom_histogram() +
  ggtitle("Square-root Inverse Simpson Distribution") +
  xlab("Square-root Inverse Simpson Index") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
```

### Faith's PD

```{r faith_pd}
# Distribution -----------------------------------------------------------------
faith_distribution <- ggplot(md_avg, aes(x = faith)) +
  geom_histogram() +
  ggtitle("Faith's Phylogenetic Distance Distribution") +
  xlab("Faith's PD") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )

# Descriptive statistics -------------------------------------------------------
faith_stats <- md_avg %>%
  summarize(
    Minimum = min(faith),
    Maximum = max(faith),
    Median = median(faith),
    Mean = mean(faith),
    SD = sd(faith)
  )

# Linear model -----------------------------------------------------------------
faith_lm <- lm(faith ~ sex + breed_clean + neutered +
                 active + body_condition + diet + biotics +
                 gastro + skin + neuro + healthy + worming +
                 antibiotics + location + age_categorical + batch,
               data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(faith_lm, which = c(1, 2, 4))

# Log Transformation -----------------------------------------------------------
md_avg$faith_log <- log(md_avg$faith)
faith_log <- lm(faith_log ~ sex + breed_clean + neutered +
                  active + body_condition + diet + biotics +
                  gastro + skin + neuro + healthy + worming +
                  antibiotics + location + age_categorical + batch,
                data = md_avg)

# Checking model fit
par(mfrow = c(1, 3))
plot(faith_log, which = 1)
plot(faith_log, which = 2)
# Adding a 1:1 line for the Q-Q plot
abline(0, 1)
plot(faith_log, which = 4)

# Transformed distribution -----------------------------------------------------
faith_log_dist <- ggplot(md_avg, aes(x = faith_log)) +
  geom_histogram() +
  ggtitle("Log-transformed Faith PD Distribution") +
  xlab("Log-transformed Faith PD") +
  ylab("Count") +
  theme(
    axis.title.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.text.x = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5)
  )
```

### Model Summary

#### Non-parametric Tests

Non parametric tests (Kruskal-Wallis and Wilcoxon) for metrics that did not hold
model assumption and could not be transformed.

```{r alpha_non_parametric}
# Omnibus Kruskal-Wallis (Wilcoxon) --------------------------------------------
# Helper function for Kruskal-Wallis of an alpha metric against each covariate
# Takes metadata, alpha metric, vector of covariates and an adjustment method
kruskal_alpha <- function(data, metric, covariates, adjust = "BH") {

  # Initializing a results list
  results <- list()

  # Iterating through the covariates
  for (cv in covariates) {
    # Covariate as a factor, dropping unused levels
    g <- droplevels(as.factor(data[[cv]]))

    # Kruskal-Wallis - reduces to Wilcoxon rank-sum for binary factors
    ht <- kruskal.test(data[[metric]] ~ g)

    # Tidy results
    results[[cv]] <- data.frame(
      covariate = cv,
      n_levels = nlevels(g),
      chi_sq = unname(ht$statistic),
      df = unname(ht$parameter),
      p = ht$p.value
    )
  }

  # Combining and BH-adjusting across the covariates
  out <- do.call(rbind, results)
  out$p_adj <- p.adjust(out$p, method = adjust)
  out <- out[order(out$p_adj), ]
  rownames(out) <- NULL
  out
}

# Pairwise comparisons - Dunn's test -------------------------------------------
# Effect size according to:
# https://stats.stackexchange.com/questions/612981/
# And: https://stats.stackexchange.com/questions/179971/
# Helper function for pairwise post-hoc for a metric for a significant
# multi-level factor in omnibus or simple effect sizes for binary factors
# Takes metadata, alpha metric, factor name, and an adjustment method
dunn_alpha <- function(data, metric, factor_name, adjust = "BH") {

  # Factor as a factor, dropping unused levels
  g <- droplevels(as.factor(data[[factor_name]]))
  y <- data[[metric]]
  lvls <- levels(g)

  # Effect size and CI for binary covariates (no actual Dunn's test)
  if (length(lvls) == 2) {
    kt <- kruskal.test(y ~ g)
    cd <- effsize::cliff.delta(y ~ g)
    return(data.frame(
      metric = metric,
      term = factor_name,
      comparison = paste(lvls, collapse = " - "),
      Z = NA_real_,
      p = kt$p.value,

      # Single comparison, no correction
      p_adj = kt$p.value,
      cliff_delta = as.numeric(cd$estimate),
      ci_lower = as.numeric(cd$conf.int[1]),
      ci_upper = as.numeric(cd$conf.int[2]),
      row.names = NULL
    ))
  }

  # Dunn's test, BH-adjusted within factor for multi-level factors
  dt <- FSA::dunnTest(data[[metric]] ~ g, method = NULL)

  # Output
  out <- dt$res

  # BH correction
  p_adj <- p.adjust(out$P.unadj, method = adjust)

  # Effect size (Cliff's delta) + CI per comparison
  eff <- do.call(rbind, lapply(out$Comparison, function(cmp) {
    lv <- trimws(strsplit(cmp, " - ")[[1]])
    keep <- g %in% lv
    cd <- effsize::cliff.delta(y[keep] ~ droplevels(g[keep]))
    data.frame(cliff_delta = as.numeric(cd$estimate),
               ci_lower = as.numeric(cd$conf.int[1]),
               ci_upper = as.numeric(cd$conf.int[2]))
  }))

  # Tidy output
  data.frame(
    metric = metric,
    term = factor_name,
    comparison = out$Comparison,
    Z = out$Z,
    p = out$P.unadj,
    p_adj = p.adjust(out$P.unadj, method = adjust),
    eff,
    row.names  = NULL
  )
}

# Covariates -------------------------------------------------------------------
covariates <- c("sex", "breed_clean", "neutered", "active",
                "body_condition", "diet", "biotics", "gastro",
                "skin", "neuro", "healthy", "worming",
                "antibiotics", "location", "age_categorical")

# Observed ---------------------------------------------------------------------
# Biological covariates
kw_observed <- kruskal_alpha(md_avg, "observed", covariates)

# Batch effect
kw_observed_batch <- kruskal_alpha(md_avg, "observed", "batch")

# Pairwise comparisons
dunn_observed_diet <- dunn_alpha(md_avg, "observed", "diet")
dunn_observed_breed <- dunn_alpha(md_avg, "observed", "breed_clean")
dunn_observed_neutered <- dunn_alpha(md_avg, "observed", "neutered")
dunn_observed_age <- dunn_alpha(md_avg, "observed", "age_categorical")
dunn_observed_active <- dunn_alpha(md_avg, "observed", "active")

# Saving results
post_observed <- list(diet = dunn_observed_diet,
                      breed = dunn_observed_breed,
                      neutered = dunn_observed_neutered,
                      age = dunn_observed_age,
                      active = dunn_observed_active)
saveRDS(object = post_observed, file = "alpha_stats_otu_set2/post_observed.RDS")

# Chao1 ------------------------------------------------------------------------
# Excluded from final analysis
# Biological covariates
kw_chao1 <- kruskal_alpha(md_avg, "chao1", covariates)

# Batch effect
kw_chao1_batch <- kruskal_alpha(md_avg, "chao1", "batch")

# Pairwise comparisons
dunn_chao1_neutered <- dunn_alpha(md_avg, "chao1", "neutered")
dunn_chao1_diet <- dunn_alpha(md_avg, "chao1", "diet")
dunn_chao1_breed <- dunn_alpha(md_avg, "chao1", "breed_clean")
dunn_chao1_active <- dunn_alpha(md_avg, "chao1", "active")
dunn_chao1_age <- dunn_alpha(md_avg, "chao1", "age_categorical")

# Saving results
post_chao1 <- list(diet = dunn_chao1_diet,
                   breed = dunn_chao1_breed,
                   neutered = dunn_chao1_neutered,
                   age = dunn_chao1_age,
                   active = dunn_chao1_active)
saveRDS(object = post_chao1, file = "alpha_stats_otu_set2/post_chao1.RDS")

# Faith's PD -------------------------------------------------------------------
# Biological covariates
kw_faith <- kruskal_alpha(md_avg, "faith", covariates)

# Batch effect
kw_faith_batch <- kruskal_alpha(md_avg, "faith", "batch")

# Pairwise comparisons
dunn_faith_diet <- dunn_alpha(md_avg, "faith", "diet")
dunn_faith_neutered <- dunn_alpha(md_avg, "faith", "neutered")
dunn_faith_breed <- dunn_alpha(md_avg, "faith", "breed_clean")
dunn_faith_active <- dunn_alpha(md_avg, "faith", "active")
dunn_faith_age <- dunn_alpha(md_avg, "faith", "age_categorical")
dunn_faith_location <- dunn_alpha(md_avg, "faith", "location")

# Saving results
post_faith <- list(diet = dunn_faith_diet,
                   breed = dunn_faith_breed,
                   neutered = dunn_faith_neutered,
                   age = dunn_faith_age,
                   active = dunn_faith_active,
                   location = dunn_faith_location)
saveRDS(object = post_faith, file = "alpha_stats_otu_set2/post_faith.RDS")
```

#### Linear Models
Summary statistics across linear models. 

```{r alpha_models}
# Anova type II for models w/o interactions
# From:
# https://www.r-bloggers.com/2011/03/anova-%E2%80%93-type-iiiiii-ss-explained/

# Helper function for post hoc comparisons for significant variables -----------
# Help from:
# https://stats.stackexchange.com/questions/102857/post-hoc-after-glm
# https://cran.r-project.org/web/packages/emmeans/vignettes/AQuickStart.html
emm_posthoc <- function(fit,
                        factors,
                        adjust = "BH") {
  mf <- model.frame(fit)

  # Factor variables in the model
  factor_vars <- names(mf)[vapply(mf, is.factor, logical(1))]

  out <- lapply(factors, function(term) {

    # Simplifying calculation grid
    # Keeping other covariates averaged across levels, proportionally-weighted
    nuisance_vars <- setdiff(factor_vars, term)
    emm <- emmeans(
      fit,
      as.formula(paste("~", term)),
      nuisance = nuisance_vars,
      weights = "proportional",
      type = "response"
    )

    # Pairwise comparisons
    prs <- pairs(emm, adjust = adjust, type = "response")

    df <- as.data.frame(summary(prs, infer = TRUE))
    df$term <- term
    df
  })

  do.call(rbind, out)
}

# Shannon - linear model -------------------------------------------------------
car::Anova(shannon_lm)

# Post-hoc for significant factors
# Breed was also significant, but may inflate the results
sig_shannon <- c("antibiotics", "age_categorical", "diet",
                 "sex", "skin")
post_shannon <- emm_posthoc(fit = shannon_lm,
                            factors = sig_shannon,
                            adjust = "BH")
saveRDS(object = post_shannon, file = "alpha_stats_otu_set2/post_shannon.RDS")

# Inverse Simpson - sqrt transformation ----------------------------------------
car::Anova(simpson_sqrt)

# Post-hoc for significant factors
# Breed was also significant, but may inflate the results
sig_simpson <- c("antibiotics", "age_categorical",
                 "sex", "skin")
post_simpson <- emm_posthoc(fit = simpson_sqrt,
                            factors = sig_simpson,
                            adjust = "BH")
saveRDS(object = post_simpson, file = "alpha_stats_otu_set2/post_simpson.RDS")
```

## Beta Diversity

Second assessment of host factors associations with the microbiome is performed
with beta diversity metrics (between-sample diversity). PERMANOVA tests are
performed, and dispersion is estimated with PERMDISP.

```{r avg_beta}
# Getting averaged metrics and saving objects ----------------------------------
# From:
# https://bioconductor.posit.co/packages/3.23/bioc/manuals/mia/man/mia.pdf
# 100 iterations, four metrics

# Bray-Curtis
if (file.exists("avg_results_otu_set2/avg_bray.RDS")) {
  avg_bray <- readRDS("avg_results_otu_set2/avg_bray.RDS")
} else {
  set.seed(122863)
  avg_bray <- getDissimilarity(
    tse,
    method = "bray",
    assay.type = "counts",
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_bray,
          file = "avg_results_otu_set2/avg_bray.RDS")
}

# Jaccard
if (file.exists("avg_results_otu_set2/avg_jac.RDS")) {
  avg_jac <- readRDS("avg_results_otu_set2/avg_jac.RDS")
} else {
  set.seed(122863)
  avg_jac <- getDissimilarity(
    tse,
    method = "jaccard",
    assay.type = "counts",
    binary = TRUE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_jac,
          file = "avg_results_otu_set2/avg_jac.RDS")
}

# Unweighted UniFrac
if (file.exists("avg_results_otu_set2/avg_uwu.RDS")) {
  avg_uwu <- readRDS("avg_results_otu_set2/avg_uwu.RDS")
} else {
  set.seed(122863)
  avg_uwu <- getDissimilarity(
    tse, method = "unifrac",
    name = "uwu",
    assay.type = "counts",
    weighted = FALSE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_uwu,
          file = "avg_results_otu_set2/avg_uwu.RDS")
}

# Weighted UniFrac
if (file.exists("avg_results_otu_set2/avg_wu.RDS")) {
  avg_wu <- readRDS("avg_results_otu_set2/avg_wu.RDS")
} else {
  set.seed(122863)
  avg_wu <- getDissimilarity(
    tse, method = "unifrac",
    name = "wu",
    assay.type = "counts",
    weighted = TRUE,
    niter = 100,
    sample = global_min
  )

  saveRDS(object = avg_wu,
          file = "avg_results_otu_set2/avg_wu.RDS")
}

# Helper function to run dispersion --------------------------------------------
# Inspired by: https://www.youtube.com/watch?v=oLf0EpMJ4yA
run_dispersion <- function(dist, group, permutations = 999) {

  bd <- betadisper(dist, group = group)
  pt <- permutest(bd, permutations = permutations)

  list(betadisper = bd, permutest = pt, p_global = pt$tab["Groups", "Pr(>F)"])
}

# Helper function for pairwise PERMANOVA comparisons ---------------------------
# Adapted from:
# https://gist.github.com/mcgoodman/58c9d1257fd1625954a4ffa1c3301939
# And: the pairwise.adonis2() function
# Using ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
pairwise_permanova <- function(dist, factor_name, covars = NULL, meta,
                               permutations = 999, adjust = "BH",
                               seed = 122863) {

  # Aligning metadata rows to the distance object
  meta <- meta[labels(dist), , drop = FALSE]
  dmat <- as.matrix(dist)

  # Factor vector and all-level combinations
  fac <- droplevels(as.factor(meta[[factor_name]]))
  pairs <- combn(levels(fac), 2, simplify = FALSE)

  # Initializing a results list
  results <- list()

  # Iterating through the possible combinations, calculating the statistics
  # for each variable restricted to a pair of values
  for (pr in pairs) {

    # Subsetting to the two levels being compared
    keep <- fac %in% pr
    d_sub <- as.dist(dmat[keep, keep])
    m_sub <- droplevels(meta[keep, , drop = FALSE])

    # Keeping only covariates that vary within the pair (more than one level)
    good_covars <- covars[sapply(covars, function(cv) {
      nlevels(droplevels(factor(m_sub[[cv]]))) > 1
    })]

    # Building formula
    rhs <- paste(c(good_covars, factor_name), collapse = " + ")
    form <- as.formula(paste("d_sub ~", rhs))

    # Marginal test w/ adjustment for covariates
    set.seed(seed)
    fit <- adonis2(form, data = m_sub, by = "margin",
                   permutations = permutations)

    # Tidy results
    results[[paste(pr, collapse = "_vs_")]] <- data.frame(
      comparison = paste(pr, collapse = "_vs_"),
      n1 = sum(m_sub[[factor_name]] == pr[1]),
      n2 = sum(m_sub[[factor_name]] == pr[2]),
      R2 = fit[factor_name, "R2"],
      F = fit[factor_name, "F"],
      p = fit[factor_name, "Pr(>F)"]
    )
  }

  # Combining and BH-adjustment within current factor
  out <- do.call(rbind, results)
  out$p_adj <- p.adjust(out$p, method = adjust)
  rownames(out) <- NULL
  out
}

# Helper function for pairwise PERMDISP comparisons ----------------------------

# Adapted from:
# https://gist.github.com/mcgoodman/58c9d1257fd1625954a4ffa1c3301939
# And: the function pairwise.adonis2()
# Using ELM (Claude Opus 4.8), https://elm.edina.ac.uk/elm/elm
pairwise_permdisp <- function(dist, factor_name, meta,
                              permutations = 999, adjust = "BH",
                              seed = 122863) {

  # Aligning metadata rows to the distance object
  meta <- meta[labels(dist), , drop = FALSE]
  dmat <- as.matrix(dist)

  # Factor vector and all level combination
  fac <- droplevels(as.factor(meta[[factor_name]]))
  pairs <- combn(levels(fac), 2, simplify = FALSE)

  # Initializing a results list
  results <- list()

  # Iterating through the possible combinations, calculating the statistics
  # for each variable restricted to a pair of values
  for (pr in pairs) {

    # Subsetting to the two levels being compared
    keep  <- fac %in% pr
    d_sub <- as.dist(dmat[keep, keep])
    g_sub <- droplevels(fac[keep])

    # Dispersion test - no covariates
    set.seed(seed)
    bd <- betadisper(d_sub, g_sub)
    pt <- permutest(bd, permutations = permutations)

    # Tidy results
    results[[paste(pr, collapse = "_vs_")]] <- data.frame(
      comparison = paste(pr, collapse = "_vs_"),
      n1 = sum(g_sub == pr[1]),
      n2 = sum(g_sub == pr[2]),
      F  = pt$tab["Groups", "F"],
      p  = pt$tab["Groups", "Pr(>F)"]
    )
  }

  # Combining and BH-adjustment within current factor
  out <- do.call(rbind, results)
  out$p_adj <- p.adjust(out$p, method = adjust)
  rownames(out) <- NULL
  out
}
```

### Bray-Curtis Dissimilarity

```{r bray_curtis}
# PERMANOVA
if (file.exists("avg_results_otu_set2/permanova_bray.RDS")) {
  permanova_bray <- readRDS("avg_results_otu_set2/permanova_bray.RDS")
} else {
  set.seed(122863)
  permanova_bray <- adonis2(avg_bray ~ sex + breed_clean + neutered +
                              active + body_condition + diet + biotics +
                              gastro + skin + neuro + healthy + worming +
                              antibiotics + location + age_categorical + batch,
                            data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_bray,
          file = "avg_results_otu_set2/permanova_bray.RDS")
}

# PERMDISP
if (file.exists("avg_results_otu_set2/disp_bray.RDS")) {
  disp_bray <- readRDS("avg_results_otu_set2/disp_bray.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_bray), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_bray), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_bray <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_bray,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_bray, file = "avg_results_otu_set2/disp_bray.RDS")
}
```

### Jaccard Distance

```{r jaccard}
# PERMANOVA
if (file.exists("avg_results_otu_set2/permanova_jac.RDS")) {
  permanova_jac <- readRDS("avg_results_otu_set2/permanova_jac.RDS")
} else {
  set.seed(122863)
  permanova_jac <- adonis2(avg_jac ~ sex + breed_clean + neutered +
                             active + body_condition + diet + biotics +
                             gastro + skin + neuro + healthy + worming +
                             antibiotics + location + age_categorical + batch,
                           data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_jac,
          file = "avg_results_otu_set2/permanova_jac.RDS")
}

# PERMDISP
if (file.exists("avg_results_otu_set2/disp_jac.RDS")) {
  disp_jac <- readRDS("avg_results_otu_set2/disp_jac.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_jac), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_jac), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_jac <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_jac,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_jac, file = "avg_results_otu_set2/disp_jac.RDS")
}
```

### Unweighted UniFrace

```{r uwu}
# PERMANOVA
if (file.exists("avg_results_otu_set2/permanova_uwu.RDS")) {
  permanova_uwu <- readRDS("avg_results_otu_set2/permanova_uwu.RDS")
} else {
  set.seed(122863)
  permanova_uwu <- adonis2(avg_uwu ~ sex + breed_clean + neutered +
                             active + body_condition + diet + biotics +
                             gastro + skin + neuro + healthy + worming +
                             antibiotics + location + age_categorical + batch,
                           data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_uwu,
          file = "avg_results_otu_set2/permanova_uwu.RDS")
}

# PERMDISP
if (file.exists("avg_results_otu_set2/disp_uwu.RDS")) {
  disp_uwu <- readRDS("avg_results_otu_set2/disp_uwu.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_uwu), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_uwu), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_uwu <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_uwu,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_uwu, file = "avg_results_otu_set2/disp_uwu.RDS")
}
```

### Weighted UniFrac

```{r wu}
# PERMANOVA
if (file.exists("avg_results_otu_set2/permanova_wu.RDS")) {
  permanova_wu <- readRDS("avg_results_otu_set2/permanova_wu.RDS")
} else {
  set.seed(122863)
  permanova_wu <- adonis2(avg_wu ~ sex + breed_clean + neutered +
                            active + body_condition + diet + biotics +
                            gastro + skin + neuro + healthy + worming +
                            antibiotics + location + age_categorical + batch,
                          data = md_avg, permutations = 999, by = "margin")

  saveRDS(object = permanova_wu, file = "avg_results_otu_set2/permanova_wu.RDS")
}

# PERMDISP
if (file.exists("avg_results_otu_set2/disp_wu.RDS")) {
  disp_wu <- readRDS("avg_results_otu_set2/disp_wu.RDS")
} else {

  # Relevant factors
  md_avg <- md_avg[labels(avg_wu), , drop = FALSE]
  perm_terms <- setdiff(rownames(permanova_wu), c("Residual", "Total"))
  fac_terms <- perm_terms[
    perm_terms %in% names(md_avg) &
      vapply(md_avg[, perm_terms, drop = FALSE], is.factor, logical(1))
  ]

  # Dispersion for relevant parameters
  disp_wu <- md_avg %>%
    select(all_of(fac_terms)) %>%
    summarize(
      across(everything(), ~ list(run_dispersion(avg_wu,
                                                 .x,
                                                 permutations = 999)))
    ) %>%
    pivot_longer(everything(), names_to = "var", values_to = "res") %>%
    mutate(
      p_disp = map_dbl(res, "p_global"),
      FDR_BH = p.adjust(p_disp, method = "BH")
    ) %>%
    select(var, p_disp, FDR_BH)

  saveRDS(object = disp_wu, file = "avg_results_otu_set2/disp_wu.RDS")
}
```

### Post-hoc Comparisons

Post-hoc pairwise comparisons for the significant multi-level factors.  
Subsetting the data for each pair combination, and performing a PERMANOVA
while controlling for the strongest factors.

```{r beta_posthoc}
# PERMANOVA pairwise comparisons -----------------------------------------------
if (file.exists("avg_results_otu_set2/pw_permanova.RDS")) {
  pw_permanova <- readRDS("avg_results_otu_set2/pw_permanova.RDS")
} else {

  # Helper function to find significant multi-level factors in a given omnibus
  # PERMANOVA
  sig_multilevel <- function(permanova, meta, alpha = 0.05) {
    terms <- setdiff(rownames(permanova), c("Residual", "Total"))

    # Keeping factors present in metadata
    terms <- terms[terms %in% names(meta) &
                     vapply(meta[terms], is.factor, logical(1))]

    # Significant in the omnibus
    pvals <- permanova[terms, "Pr(>F)"]
    terms <- terms[!is.na(pvals) & pvals < alpha]
    # Only w/ more than two levels
    terms[vapply(terms,
                 function(v) nlevels(droplevels(meta[[v]])) > 2,
                 logical(1))]
  }

  # Confounders
  adj <- c("batch", "age_categorical", "antibiotics", "diet")

  # Helper function to run all significant multi-level factors for one metric
  run_metric <- function(dist, permanova) {
    meta_m <- md_avg[labels(dist), , drop = FALSE]
    facs <- sig_multilevel(permanova, meta_m)
    if (!length(facs)) return(NULL)

    res <- lapply(facs, function(f) {
      out <- pairwise_permanova(
        dist = dist,
        factor_name = f,
        # Not adjusting a factor for itself
        covars = setdiff(adj, f),
        meta = md_avg
      )
      out$term <- f
      out
    })
    do.call(rbind, res)
  }

  set.seed(122863)
  pw_permanova <- list(
    bray = run_metric(avg_bray, permanova_bray),
    jac = run_metric(avg_jac, permanova_jac),
    uwu = run_metric(avg_uwu, permanova_uwu),
    wu = run_metric(avg_wu, permanova_wu)
  )

  saveRDS(pw_permanova, "avg_results_otu_set2/pw_permanova.RDS")
}

# PERMDISP pairwise comparisons ------------------------------------------------
if (file.exists("avg_results_otu_set2/pw_permdisp.RDS")) {
  pw_permdisp <- readRDS("avg_results_otu_set2/pw_permdisp.RDS")
} else {

  # Helper function to find significant multi-level factors in a given omnibus
  # PERMANOVA
  sig_multilevel <- function(permanova, meta, alpha = 0.05) {
    terms <- setdiff(rownames(permanova), c("Residual", "Total"))

    # Keeping factors present in metadata
    terms <- terms[terms %in% names(meta) &
                     vapply(meta[terms], is.factor, logical(1))]

    # Significant in the omnibus
    pvals <- permanova[terms, "Pr(>F)"]
    terms <- terms[!is.na(pvals) & pvals < alpha]

    # Only w/ more than two levels
    terms[vapply(terms,
                 function(v) nlevels(droplevels(meta[[v]])) > 2,
                 logical(1))]
  }

  # Helper function to run all significant multi-level factors for one metric
  # (dispersion takes no covariates)
  disp_metric <- function(dist, permanova) {
    meta_m <- md_avg[labels(dist), , drop = FALSE]
    facs <- sig_multilevel(permanova, meta_m)
    if (!length(facs)) return(NULL)

    res <- lapply(facs, function(f) {
      out <- pairwise_permdisp(
        dist = dist,
        factor_name = f,
        meta = md_avg
      )
      out$term <- f
      out
    })
    do.call(rbind, res)
  }

  set.seed(122863)
  pw_permdisp <- list(
    bray = disp_metric(avg_bray, permanova_bray),
    jac = disp_metric(avg_jac, permanova_jac),
    uwu = disp_metric(avg_uwu, permanova_uwu),
    wu = disp_metric(avg_wu, permanova_wu)
  )

  saveRDS(pw_permdisp, "avg_results_otu_set2/pw_permdisp.RDS")
}
```

## Taxonomy and Abundance

Extracting relative abundance and calculating occurrence by taxonomic ranks.

```{r abundance}
# Relative abundance
ps_rel <- microbiome::transform(ps_f, "compositional")
taxon_table <- psmelt(ps_rel)
```

### Phylum

```{r phylum}
# Retrieving phylum-level information
# Mean taxa across samples
tax_table_phy <- taxon_table %>%

  # Handling missing data
  mutate(
    Phylum = as.character(Phylum),
    Phylum = trimws(Phylum),
    Phylum = na_if(Phylum, ""),
    Phylum = replace_na(Phylum, "Unclassified")
  ) %>%

  # Phylum-abundance per sample
  group_by(Sample, Phylum) %>%
  summarize(sample_phylum_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of phyla across samples
  group_by(Phylum) %>%
  summarize(

    # Number of samples with phyla
    num_samples = sum(sample_phylum_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_phylum_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_phylum_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_phy$num_samples == 1)

# Top phyla and unclassified
head(tax_table_phy)
tax_table_phy[tax_table_phy$Phylum == "Unclassified", ]
```

### Class

```{r class}
# Retrieving class-level information
# Mean taxa across samples
tax_table_class <- taxon_table %>%

  # Handling missing data
  mutate(
    Class = as.character(Class),
    Class = trimws(Class),
    Class = na_if(Class, ""),
    Class = replace_na(Class, "Unclassified")
  ) %>%

  # Class-abundance per sample
  group_by(Sample, Class) %>%
  summarize(sample_class_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of class across samples
  group_by(Class) %>%
  summarize(

    # Number of samples with class
    num_samples = sum(sample_class_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_class_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_class_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_class$num_samples == 1)

# Top classes and unclassified
head(tax_table_class)
tax_table_class[tax_table_class$Class == "Unclassified", ]
```

### Order

```{r order}
# Retrieving order-level information
# Mean taxa across samples
tax_table_order <- taxon_table %>%

  # Handling missing data
  mutate(
    Order = as.character(Order),
    Order = trimws(Order),
    Order = na_if(Order, ""),
    Order = replace_na(Order, "Unclassified")
  ) %>%

  # Order-abundance per sample
  group_by(Sample, Order) %>%
  summarize(sample_order_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of order across samples
  group_by(Order) %>%
  summarize(

    # Number of samples with order
    num_samples = sum(sample_order_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_order_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_order_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%
  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_order$num_samples == 1)

# Top orders and unclassified
head(tax_table_order)
tax_table_order[tax_table_order$Order == "Unclassified", ]
```

### Family

```{r family}
# Retrieving family-level information
# Mean taxa across samples
tax_table_family <- taxon_table %>%

  # Handling missing data
  mutate(
    Family = as.character(Family),
    Family = trimws(Family),
    Family = na_if(Family, ""),
    Family = replace_na(Family, "Unclassified")
  ) %>%

  # Family-abundance per sample
  group_by(Sample, Family) %>%
  summarize(sample_family_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of family across samples
  group_by(Family) %>%
  summarize(

    # Number of samples with family
    num_samples = sum(sample_family_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_family_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_family_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_family$num_samples == 1)

# Top families and unclassified
head(tax_table_family)
tax_table_family[tax_table_family$Family == "Unclassified", ]
```

### Genus

```{r genus}
# Retrieving genus-level information
# Mean taxa across samples
tax_table_genus <- taxon_table %>%

  # Handling missing data
  mutate(
    Genus = as.character(Genus),
    Genus = trimws(Genus),
    Genus = na_if(Genus, ""),
    Genus = replace_na(Genus, "Unclassified")
  ) %>%

  # Genus-abundance per sample
  group_by(Sample, Genus) %>%
  summarize(sample_genus_abund = sum(Abundance),
            .groups = "drop") %>%

  # Total abundance of genus across samples
  group_by(Genus) %>%
  summarize(

    # Number of samples with genus
    num_samples = sum(sample_genus_abund > 0),

    # Sum of relative abundance across samples
    sum_abund = sum(sample_genus_abund),

    # Mean of relative abundance across samples
    mean_abund = mean(sample_genus_abund),
    .groups = "drop"
  ) %>%

  # Abundance in percentage
  mutate(perc_abund = mean_abund * 100) %>%

  # And descending order
  arrange(desc(perc_abund))

# Single sample
sum(tax_table_genus$num_samples == 1)

# Top genera and unclassified
head(tax_table_genus)
tax_table_genus[tax_table_genus$Genus == "Unclassified", ]
```

### Differential Abundance

Not included in the final analysis, partly due to very high amount of
retained taxa after prevalence filtering.

```{r prevalent_abundances}
# Checking prevalence threshold
# Helper function to calculate prevalence
get_prevalence <- function(taxon) {
  prevalence <- sum(taxon > 0) / length(taxon)
  prevalence
}

# Non-rarefied data
ps_prev <- filter_taxa(ps_f,
                       function(taxon) get_prevalence(taxon) > 0.1,
                       prune = TRUE)

# Some checks
nsamples(ps_prev)
ntaxa(ps_prev) # 672 taxa remaining
summary(sample_sums(ps_prev))
```
